---
title: "minicondition"
---

```{r}
#| include: false
source(file.path("minicondition", "minicondition.R"))
```

`minicondition` provides `mcond_abort()`/`_warn()`/`_inform()` -- thin
wrappers around base R's own condition system that signal a *classed*
condition, catchable specifically by that class -- plus `mcond_assert()`
built on top of `mcond_abort()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minicondition)
for the rlang-parity notes; this page focuses on the catch-by-class
behaviour, since that's the main reason to reach for `minicondition`
over plain `stop()`/`warning()`/`message()`, and finishes with
`mcond_assert()`'s handling of `NA`.

## Functions

- `mcond_abort(message, class = NULL)` -- signal a classed error.
- `mcond_warn(message, class = NULL)` -- signal a classed warning.
- `mcond_inform(message, class = NULL)` -- signal a classed message.
- `mcond_assert(expr, message, class = NULL)` -- call `mcond_abort()`
  if `expr` isn't entirely `TRUE`; `NA` counts as a failure.

`minicondition` has one internal helper, `.mcond_condition()`, which
builds the classed condition object that `mcond_abort()`/`_warn()`/
`_inform()` each hand off to base `stop()`/`warning()`/`message()`.

## Why classes matter

A plain `stop("bad input")` produces a condition of class `c("simpleError",
"error", "condition")` -- generic, and indistinguishable from any other
error at the point where it's caught. That's fine for a one-off script,
but it's a problem inside a package: callers who want to react to
*this specific kind* of failure (retry, fall back to a default, show a
custom message) have no reliable way to distinguish it from every other
error your code might raise. `mcond_abort()` solves that by adding an
extra class on top of `"error"`, so `tryCatch()` can dispatch on it
directly:

```{r}
tryCatch(
  mcond_abort("bad input", class = "mypkg_invalid_input"),
  mypkg_invalid_input = function(e) message("caught: ", conditionMessage(e))
)
```

Note that `mcond_abort()` uses base `stop()` internally, not
`rlang::abort()` -- but the mechanics of catching by class are
identical either way. `tryCatch()`/`withCallingHandlers()` dispatch on
a condition object's `class` attribute, regardless of what function
signalled it:

```{r}
e <- tryCatch(mcond_abort("x", class = "mypkg_invalid_input"), error = function(e) e)
class(e)
```

## Unmatched handlers fall through; `error` still catches it

A handler for an unrelated class simply doesn't match -- `tryCatch()`
tries each handler in order and uses the first one whose class matches.
Since the condition is still, underneath the custom class, an ordinary
`error` (see the `class(e)` output above), a generic `error` handler
still catches it as a fallback if nothing more specific matches first:

```{r}
tryCatch(
  mcond_abort("bad input", class = "mypkg_invalid_input"),
  mypkg_other_class = function(e) message("this handler never runs"),
  error = function(e) message("caught generically as a plain error: ", conditionMessage(e))
)
```

This is the same layering `rlang` conditions use: a custom class sits
*in addition to* the base `error`/`warning`/`message` class, not
instead of it, so code that only knows about the generic type still
works.

## Warnings and messages work the same way

`mcond_warn()` signals a classed warning, catchable via
`withCallingHandlers()` (warnings, unlike errors, don't unwind the
call stack, so catching them with `tryCatch()` would abandon whatever
code was about to run after the warning -- `withCallingHandlers()`
lets the handler run and then hand control back):

```{r}
withCallingHandlers(
  mcond_warn("using a fallback default", class = "mypkg_fallback_used"),
  mypkg_fallback_used = function(w) {
    message("caught warning: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

The `invokeRestart("muffleWarning")` call suppresses the warning from
also being printed by R's default warning handler -- without it, the
warning would print a second time after the custom handler runs.

`mcond_inform()` signals a classed message -- uncaught, it just prints
like an ordinary `message()` call, since messages are typically meant
to be seen unless something specifically wants to intercept them:

```{r}
mcond_inform("starting up", class = "mypkg_startup")
```

## `mcond_assert()`: abort if a condition doesn't hold

`mcond_assert()` is `mcond_abort()` plus a check: pass it a logical
expression, and it aborts (with the same classed-error mechanics as
above) only if that expression isn't entirely `TRUE`. A passing
assertion is silent, so it's safe to sprinkle through code as a
precondition check without cluttering normal output:

```{r}
x <- 5
mcond_assert(x > 0, "x must be positive.")
```

A failing one raises a classed error, catchable the same way as
`mcond_abort()` -- which makes sense, since that's exactly what it
calls internally:

```{r}
tryCatch(
  mcond_assert(x > 0 && x < 3, "x must be between 0 and 3.", class = "mypkg_range_error"),
  mypkg_range_error = function(e) message("caught: ", conditionMessage(e))
)
```

## `NA` counts as a failure, not a crash

`mcond_assert()` accepts a vector, not just a single logical value, and
checks that *every* element is `TRUE`. That raises a subtlety: what
should happen if the vector contains `NA`? This is a deliberate fix
relative to the source this was adapted from: a naive `if (any(expr ==
FALSE))` errors outright when `expr` contains `NA` ("missing value
where TRUE/FALSE needed"), which would crash the assertion check itself
rather than raising the intended, informative assertion error.
`mcond_assert()` checks `anyNA(expr)` explicitly first, so a vector
containing `NA` fails the assertion cleanly, with the message you
asked for, instead of an unrelated low-level error:

```{r}
#| error: true
y <- c(TRUE, TRUE, NA)
mcond_assert(y, "y must not contain NA.", class = "mypkg_na_error")
```
