---
title: "minicondition"
---

```{r}
#| include: false
source(file.path("minicondition", "minicondition.R"))
```

`minicondition` provides `mcond_abort()`/`_warn()`/`_inform()` -- thin
wrappers around base R's own condition system that signal a *classed*
condition, catchable specifically by that class -- plus `mcond_assert()`
built on top of `mcond_abort()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minicondition)
for the rlang-parity notes; this page focuses on the catch-by-class
behaviour and `mcond_assert()`'s `NA` handling.

## Catching a classed error by its class

`tryCatch()` dispatches on a condition's class, whether that condition
was signalled via `rlang::abort()` or, as here, via a plain classed
condition object passed to base `stop()`:

```{r}
tryCatch(
  mcond_abort("bad input", class = "mypkg_invalid_input"),
  mypkg_invalid_input = function(e) message("caught: ", conditionMessage(e))
)
```

A handler for an unrelated class doesn't match -- the condition is
still an ordinary `error` underneath, so a generic `error` handler
catches it as a fallback:

```{r}
tryCatch(
  mcond_abort("bad input", class = "mypkg_invalid_input"),
  mypkg_other_class = function(e) message("this handler never runs"),
  error = function(e) message("caught generically as a plain error: ", conditionMessage(e))
)
```

## Warnings and messages work the same way

`mcond_warn()` signals a classed warning, catchable via
`withCallingHandlers()`:

```{r}
withCallingHandlers(
  mcond_warn("using a fallback default", class = "mypkg_fallback_used"),
  mypkg_fallback_used = function(w) {
    message("caught warning: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

`mcond_inform()` signals a classed message -- uncaught, it just prints
like an ordinary `message()` call:

```{r}
mcond_inform("starting up", class = "mypkg_startup")
```

## `mcond_assert()`: abort if a condition doesn't hold

A passing assertion is silent:

```{r}
x <- 5
mcond_assert(x > 0, "x must be positive.")
```

A failing one raises a classed `mcond_abort()` error, catchable the
same way:

```{r}
tryCatch(
  mcond_assert(x > 0 && x < 3, "x must be between 0 and 3.", class = "mypkg_range_error"),
  mypkg_range_error = function(e) message("caught: ", conditionMessage(e))
)
```

## `NA` counts as a failure, not a crash

This is a deliberate fix relative to the source this was adapted from:
`if (any(expr == FALSE))` errors outright when `expr` contains `NA`
("missing value where TRUE/FALSE needed"), rather than raising the
intended assertion error. `mcond_assert()` checks `anyNA(expr)`
explicitly, so a vector containing `NA` fails the assertion cleanly:

```{r}
#| error: true
y <- c(TRUE, TRUE, NA)
mcond_assert(y, "y must not contain NA.", class = "mypkg_na_error")
```
