---
title: "minitable"
---

```{r}
#| include: false
source(file.path("minitable", "minitable.R"))
```

`minitable` provides a few `tibble`-style construction/coercion
helpers -- `mtable_tibble()`, `mtable_as_tibble()`,
`mtable_rownames_to_column()`, `mtable_add_row()` -- built entirely on
`data.frame()`/`as.data.frame()`. Results are plain data frames, not
tibbles. See the
[README](https://github.com/djnavarro/minis/tree/main/minitable) for
the full scope notes; this page walks through each function, including
the cross-column-reference behaviour that's the trickiest part of the
mini to get right.

## Functions

- `mtable_tibble(...)` -- build a data frame column by column,
  sequentially, so later columns can refer to earlier ones by name.
- `mtable_as_tibble(x, ...)` -- coerce `x` to a plain data frame via
  `as.data.frame()`, preserving literal (non-syntactic) column names.
- `mtable_rownames_to_column(.data, var)` -- move `.data`'s row names
  into an explicit column named `var`; a no-op if the row names are
  just the default sequential ones.
- `mtable_add_row(.data, ...)` -- append a single row to `.data`, with
  the same cross-column-reference support as `mtable_tibble()`.

`minitable` has one internal helper, `.mtable_drop_dup_list()`, which
`mtable_tibble()` and `mtable_add_row()` both use to resolve what a
column reference means when a name is reused more than once while
constructing columns in sequence (see below).

## `mtable_tibble()`: sequential, cross-referencing construction

Building a data frame with `data.frame()` directly requires every
column to already exist as a complete vector before the call -- you
can't refer to a column you're building in the same call. Like
`tibble::tibble()`, `mtable_tibble()` lifts that restriction: later
columns can refer to earlier ones by name, because each column
expression is evaluated in order, in an environment built up from the
columns already constructed:

```{r}
mtable_tibble(a = 1:3, b = a * 2, c = a + b)
```

Here `b` refers to `a`, and `c` refers to both `a` and `b` -- neither
would resolve to anything if evaluated outside the call, since `a`,
`b`, and `c` don't otherwise exist in this session. Getting this right
took real care in the implementation: the sequential-evaluation logic
lives directly inside `mtable_tibble()` itself, rather than in a
shared helper that `...` gets forwarded through. Forwarding `...` to a
separate function turns each argument into an opaque `..1`/`..2`
pronoun that resolves free variables against the *caller's*
environment instead of the column-by-column environment being built
up -- which would silently break exactly this feature, evaluating `b`
and `c` against whatever (unrelated) `a`/`b` might exist in the
calling code, or erroring if there simply isn't one.

## Unnamed arguments are named after their deparsed expression

Just as in `tibble::tibble()`, an argument with no `name =` is named
after its own source expression, converted to a string:

```{r}
mtable_tibble(1:3, sqrt(1:3))
```

This is mostly useful for quick, throwaway construction; naming
columns explicitly (`mtable_tibble(x = 1:3, y = sqrt(1:3))`) is more
readable for anything that will be used again later.

## `mtable_as_tibble()`: coercion only

There's no tibble class or special printing here -- it's just
`as.data.frame()` with `check.names = FALSE`, so a literal,
non-syntactic column name (like `"1:3"` from the previous section)
survives unchanged instead of being sanitised by `make.names()`:

```{r}
mtable_as_tibble(list(a = 1:2, b = c("x", "y")))
```

The difference from calling `as.data.frame()` directly shows up with a
name that isn't a syntactically valid R identifier, like the `"1:3"`
name `mtable_tibble()` generated automatically two sections ago:

```{r}
names(mtable_as_tibble(list(`1:3` = 1:3)))
names(as.data.frame(list(`1:3` = 1:3)))
```

`as.data.frame()`'s default `check.names = TRUE` rewrites `"1:3"` into
the syntactically valid `"X1.3"`; `mtable_as_tibble()` passes
`check.names = FALSE` through instead, so the literal name survives.

## `mtable_rownames_to_column()`: no-op on default row names

A data frame with only the default sequential row names (`"1"`, `"2"`,
...) is returned unchanged -- there's nothing meaningful to move into
a column:

```{r}
identical(
  mtable_rownames_to_column(data.frame(x = 1:3), var = "id"),
  data.frame(x = 1:3)
)
```

Real (non-default) row names do get moved into a new column, placed
first:

```{r}
mtable_rownames_to_column(head(mtcars, 3), var = "model")
```

## `mtable_add_row()`: the same cross-referencing support

Appending a row supports the same sequential evaluation as
`mtable_tibble()`, so a later value in the new row can refer to an
earlier one -- either a column already in `.data`, or one being added
in the same call:

```{r}
df <- mtable_tibble(a = 1:2, b = a * 10)
df
mtable_add_row(df, a = 3, b = a * 10)
```

Here `a * 10` in the new row refers to the `a = 3` just supplied in
that same call, not to the existing column `a` in `df` -- the most
recently constructed value for a given name always wins, which is what
`.mtable_drop_dup_list()` is for internally.
