---
title: "minitable"
---

```{r}
#| include: false
source(file.path("minitable", "minitable.R"))
```

`minitable` provides a few `tibble`-style construction/coercion
helpers -- `mtable_tibble()`, `mtable_as_tibble()`,
`mtable_rownames_to_column()`, `mtable_add_row()` -- built entirely on
`data.frame()`/`as.data.frame()`. Results are plain data frames, not
tibbles. See the
[README](https://github.com/djnavarro/minis/tree/main/minitable) for
the full scope notes; this page walks through each function, including
the cross-column-reference behaviour that took real rework to get
right (see `.agents/HISTORY.md` in the repository for the bug that
surfaced along the way).

## `mtable_tibble()`: sequential, cross-referencing construction

Like `tibble::tibble()`, later columns can refer to earlier ones by
name -- each column is evaluated in order, in an environment built
from the columns already constructed:

```{r}
mtable_tibble(a = 1:3, b = a * 2, c = a + b)
```

This only works because the sequential-evaluation logic lives directly
inside `mtable_tibble()` itself, rather than in a shared helper that
`...` gets forwarded through -- forwarding `...` turns each argument
into an opaque `..1`/`..2` pronoun that resolves free variables against
the *caller's* environment instead of the custom one being built up
column by column, silently breaking exactly this feature.

## Unnamed arguments are named after their deparsed expression

Just as in `tibble::tibble()`, an argument with no `name =` is named
after its own source expression:

```{r}
mtable_tibble(1:3, sqrt(1:3))
```

## `mtable_as_tibble()`: coercion only

There's no tibble class or special printing here -- it's just
`as.data.frame()` with `check.names = FALSE` (so literal, non-syntactic
column names survive):

```{r}
mtable_as_tibble(list(a = 1:2, b = c("x", "y")))
```

## `mtable_rownames_to_column()`: no-op on default row names

A data frame with only the default sequential row names (`"1"`, `"2"`,
...) is returned unchanged -- there's nothing meaningful to move into
a column:

```{r}
identical(
  mtable_rownames_to_column(data.frame(x = 1:3), var = "id"),
  data.frame(x = 1:3)
)
```

Real (non-default) row names do get moved into a new column:

```{r}
mtable_rownames_to_column(head(mtcars, 3), var = "model")
```

## `mtable_add_row()`: the same cross-referencing support

Appending a row supports the same sequential evaluation as
`mtable_tibble()`, so a later value in the new row can refer to an
earlier one:

```{r}
df <- mtable_tibble(a = 1:2, b = a * 10)
mtable_add_row(df, a = 3, b = a * 10)
```
