---
title: "minitable"
---

```{r}
#| include: false
source(file.path("minitable", "minitable.R"))
```

The goal with `minitable` is to provide a few `tibble`-like construction/coercion helpers built entirely on data frames. Results are plain data frames, not actual tibbles. See the [README](https://github.com/djnavarro/minis/tree/main/minitable) for the full scope notes; this page walks through each function, including the cross-column-reference behaviour that's the trickiest part of the mini to get right.

## Functions

There are currently four core functions:

- `mtable_tibble(...)` -- build a data frame column by column, sequentially, so later columns can refer to earlier ones by name.
- `mtable_as_tibble(x, ...)` -- coerce `x` to a plain data frame via `as.data.frame()`, preserving literal (non-syntactic) column names.
- `mtable_rownames_to_column(.data, var)` -- move `.data`'s row names into an explicit column named `var`; a no-op if the row names are just the default sequential ones.
- `mtable_add_row(.data, ...)` -- append a single row to `.data`, with the same cross-column-reference support as `mtable_tibble()`.

Additionally `minitable` has one internal helper:

-  `.mtable_drop_dup_list()` is used by both `mtable_tibble()` and `mtable_add_row()` to resolve what a column reference means when a name is reused more than once while constructing columns in sequence (see below).

## Sequential, cross-referencing construction

Building a data frame with `data.frame()` directly requires every column to already exist as a complete vector before the call -- you can't refer to a column you're building in the same call. Like `tibble::tibble()`, `mtable_tibble()` lifts that restriction: later columns can refer to earlier ones by name, because each column expression is evaluated in order, in an environment built up from the columns already constructed:

```{r}
mtable_tibble(a = 1:3, b = a * 2, c = a + b)
```

Here `b` refers to `a`, and `c` refers to both `a` and `b` -- neither would resolve to anything if evaluated outside the call, since `a`, `b`, and `c` don't otherwise exist in this session. Getting this right took real care in the implementation, but you can take a look at the code if you're interested.

## Unnamed arguments are named after their deparsed expression

Just as in `tibble::tibble()`, an argument with no `name =` is named after its own source expression, converted to a string:

```{r}
mtable_tibble(1:3, sqrt(1:3))
```

This is mostly useful for quick, throwaway construction; naming columns explicitly (`mtable_tibble(x = 1:3, y = sqrt(1:3))`) is more readable for anything that will be used again later.

## Minimalist coercion

There's no tibble class or special printing here -- it's just `as.data.frame()` with `check.names = FALSE`, so a literal, non-syntactic column name (like `"1:3"` from the previous section) survives unchanged instead of being sanitised by `make.names()`:

```{r}
mtable_as_tibble(list(a = 1:2, b = c("x", "y")))
```

The difference from calling `as.data.frame()` directly shows up with a name that isn't a syntactically valid R identifier, like the `"1:3"` name `mtable_tibble()` generated automatically two sections ago:

```{r}
names(mtable_as_tibble(list(`1:3` = 1:3)))
names(as.data.frame(list(`1:3` = 1:3)))
```

`as.data.frame()`'s default `check.names = TRUE` rewrites `"1:3"` into the syntactically valid `"X1.3"`; `mtable_as_tibble()` passes `check.names = FALSE` through instead, so the literal name survives.

## Rowname to column conversion

A data frame with only the default sequential row names (`"1"`, `"2"`, ...) is returned unchanged -- there's nothing meaningful to move into a column:

```{r}
identical(
  mtable_rownames_to_column(data.frame(x = 1:3), var = "id"),
  data.frame(x = 1:3)
)
```

Real (non-default) row names do get moved into a new column, placed first:

```{r}
mtable_rownames_to_column(head(mtcars, 3), var = "model")
```

## Adding rows

Appending a row with `mtable_add_row()` supports the same sequential evaluation as `mtable_tibble()`, so a later value in the new row can refer to an earlier one: either a column already in `.data`, or one that is added in the same call.

```{r}
df <- mtable_tibble(a = 1:2, b = a * 10)
df
mtable_add_row(df, a = 3, b = a * 10)
```

Here `a * 10` in the new row refers to the `a = 3` just supplied in that same call, not to the existing column `a` in `df` -- the most recently constructed value for a given name always wins, which is what `.mtable_drop_dup_list()` is for internally.

## Scope: data frame dropping still exists

Nothing in this mini changes the fact that the return objects are genuine data frames, which means that `[` still behaves exactly the same as usual for a data frame, and subsetting is therefore subject to the data frame dropping rule. It is possible that a future extension could layer a very thin S3 class that addresses this, but I am cautious about this: the risk of introducing unexpected behaviour if a "quasi-tibble" classed object escapes into the wild seems too much to countenance in a mini. 
