---
title: "minifilter"
---

```{r}
#| include: false
source(file.path("minifilter", "minifilter.R"))
```

`minifilter` provides a single function, `mfilter_filter()`, a minimal
reimplementation of `dplyr::filter()`: it keeps rows of a data frame
where all of a set of unquoted conditions are `TRUE`. See the
[README](https://github.com/djnavarro/minis/tree/main/minifilter) for
scope notes; this page walks through its behaviour on a small toy data
frame.

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## A single condition

Columns are referred to by bare name, evaluated in the scope of the
data frame:

```{r}
mfilter_filter(fruits, in_stock)
```

## Multiple conditions combine with AND

Passing more than one condition is equivalent to combining them with
`&`:

```{r}
mfilter_filter(fruits, in_stock, price < 4)
```

## `NA` conditions drop the row

`date` has a missing `price`, so any condition involving `price` is
`NA` for that row rather than `TRUE`/`FALSE` -- and `mfilter_filter()`
drops it, matching `dplyr::filter()`'s treatment of `NA` as "not kept":

```{r}
mfilter_filter(fruits, price > 1)
```

## No conditions returns the data frame unchanged

```{r}
identical(mfilter_filter(fruits), fruits)
```

This is a deliberate fix relative to the poorman logic
`minifilter` was adapted from, which didn't guard against empty `...`
and would have silently dropped every row instead -- see the
[README](https://github.com/djnavarro/minis/tree/main/minifilter#deliberate-fix-relative-to-the-source-this-was-adapted-from)
for details.
