---
title: "minifilter"
---

```{r}
#| include: false
source(file.path("minifilter", "minifilter.R"))
```

`minifilter` provides a single function, `mfilter_filter()`, a minimal
reimplementation of `dplyr::filter()`: it keeps rows of a data frame
where all of a set of unquoted conditions are `TRUE`. See the
[README](https://github.com/djnavarro/minis/tree/main/minifilter) for
scope notes; this page walks through its behaviour on a small toy data
frame, including two behaviours worth knowing about before you rely on
it: how missing values are handled, and what happens with no
conditions at all.

## Functions

There is a single core function to this mini:

- `mfilter_filter(.data, ...)` -- keep rows of `.data` where every
  condition in `...` evaluates to `TRUE`. Conditions are evaluated in
  the scope of `.data`, so columns can be referred to by bare name.

There is also one internal helper:

-  `.mfilter_dotdotdot()` captures the unevaluated `...` conditions so each one can be evaluated against the data frame individually, rather than evaluated eagerly against the caller's environment.

## A minimal data set

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## A single condition

Columns are referred to by bare name, evaluated in the scope of the
data frame -- there's no need to write `fruits$in_stock` or wrap the
call in `with()`, as you would for
[`minicase`](../minicase/)'s `mcase_case_when()`:

```{r}
mfilter_filter(fruits, in_stock)
```

## Multiple conditions combine with AND

Passing more than one condition is equivalent to combining them with
`&`, so a row is kept only if every condition holds. This is often more
readable than a single expression joined with `&&`/`&`, especially as
the number of conditions grows:

```{r}
mfilter_filter(fruits, in_stock, price < 4)
```

That's equivalent to writing the conditions out explicitly and
combining them yourself:

```{r}
identical(
  mfilter_filter(fruits, in_stock, price < 4),
  fruits[fruits$in_stock & fruits$price < 4 & !is.na(fruits$price), ]
)
```

## `NA` conditions drop the row

`date` has a missing `price`, so any condition involving `price` is
`NA` for that row rather than `TRUE`/`FALSE`. `mfilter_filter()` treats
`NA` the same way `dplyr::filter()` does: as "not proven to be kept",
so the row is dropped rather than kept by default or raising an error:

```{r}
mfilter_filter(fruits, price > 1)
```

Notice that `date` disappears from the result entirely, rather than
appearing with `NA` values or being kept because "we don't know" --
this is a common surprise for anyone used to base R's `[` subsetting,
where an `NA` index instead produces a row of `NA`s:

```{r}
fruits[fruits$price > 1, ]
```

## No conditions returns the data frame unchanged

Calling `mfilter_filter()` with no conditions at all is a legitimate,
supported no-op -- useful when conditions are built up programmatically
and might end up empty:

```{r}
identical(mfilter_filter(fruits), fruits)
```

This is a deliberate fix relative to the poorman logic `minifilter` was
adapted from, which didn't guard against empty `...` and would have
silently dropped every row instead -- see the
[README](https://github.com/djnavarro/minis/tree/main/minifilter#deliberate-fix-relative-to-the-source-this-was-adapted-from)
for details.
