---
title: "minirx"
---

```{r}
#| include: false
source(file.path("minirx", "minirx.R"))
```

The purpose of **minirx** is to provide a zero-dependency
reimplementation of stringr's regex-based pattern-matching verbs from
the [**stringr**](https://stringr.tidyverse.org/) package: detect,
extract, match, replace, remove, split, count, and locate. Every
function takes a single PCRE pattern, applied via base R's own
`perl = TRUE` regex engine rather than stringr/stringi's ICU engine.
The [README](https://github.com/djnavarro/minis/tree/main/minirx)
provides scope notes; this page walks through each function on a
small set of toy strings, including two base-R quirks this mini
deliberately works around: `grepl()` silently turning a missing string
into `FALSE` instead of `NA`, and `regmatches()` silently dropping
non-matching elements instead of leaving a hole. (For non-regex string
manipulation like padding/trimming/case conversion, see the companion
[**ministr**](../ministr/) vignette.)

## Functions

The [**minirx.R**](https://github.com/djnavarro/minis/blob/main/minirx/minirx.R)
script supplies fifteen core functions (grouped into eleven bullets below, since a few come in matched pairs):

- `.rx_detect(string, pattern, negate)` is used to detect whether a
  pattern matches.
- `.rx_starts(string, pattern, negate)` and
  `.rx_ends(string, pattern, negate)` are used to detect whether a
  string starts/ends with a pattern.
- `.rx_extract(string, pattern)` is used to extract the first match.
- `.rx_extract_all(string, pattern)` is used to extract every match.
- `.rx_match(string, pattern)` is used to extract the first match plus
  its capture groups, as a matrix.
- `.rx_match_all(string, pattern)` is used to extract every match plus
  its capture groups, as a list of matrices.
- `.rx_replace(string, pattern, replacement)` and
  `.rx_replace_all(string, pattern, replacement)` are used to replace
  the first/every match.
- `.rx_remove(string, pattern)` and `.rx_remove_all(string, pattern)`
  are used to remove the first/every match.
- `.rx_split(string, pattern)` is used to split a string on every
  match.
- `.rx_count(string, pattern)` is used to count non-overlapping
  matches.
- `.rx_locate(string, pattern)` and `.rx_locate_all(string, pattern)`
  are used to locate the position(s) of the first/every match.

It contains no internal helpers.

## Detecting matches

The `.rx_detect()` function reports whether a pattern matches anywhere
in each string. A missing string reports `NA` rather than `FALSE`,
unlike calling base `grepl()` directly:

```{r}
orders <- c("order-123", "ORDER-456", "no id here", NA)
.rx_detect(orders, "[0-9]+")
grepl("[0-9]+", orders)
```

The `negate` argument flips the result (still leaving `NA` alone), and
`.rx_starts()`/`.rx_ends()` anchor the same check to the start or end
of the string:

```{r}
.rx_detect(orders, "[0-9]+", negate = TRUE)
.rx_starts(orders, "order", negate = FALSE)
.rx_ends(orders, "[0-9]+")
```

Note that `.rx_starts()` treats case literally, the same way
`stringr::str_starts()` does -- `"ORDER-456"` doesn't start with
`"order"` above. Passing `(?i)` at the front of the pattern, PCRE's
inline case-insensitivity flag, switches that off:

```{r}
.rx_starts(orders, "(?i)order")
```

## Extracting matches

The `.rx_extract()` function extracts the first match from each
string, filling in `NA` where there's no match rather than shortening
the result -- a fix relative to calling
`regmatches(string, regexpr(pattern, string, perl = TRUE))` directly,
which drops non-matching strings from the output entirely:

```{r}
.rx_extract(orders, "[0-9]+")
```

The `.rx_extract_all()` function returns every match per string, as a
list:

```{r}
.rx_extract_all(c("a1 b2 c3", "no digits", NA), "[0-9]")
```

## Matching with capture groups

The `.rx_match()` function is like `.rx_extract()`, but also splits
out each capture group into its own column:

```{r}
.rx_match(orders, "([a-zA-Z]+)-([0-9]+)")
```

The `.rx_match_all()` function does the same for every match in a
string, returning one matrix per input:

```{r}
.rx_match_all(c("a1 b2", "none", NA), "([a-z])([0-9])")
```

## Replacing and removing matches

The `.rx_replace()`/`.rx_replace_all()` functions replace the
first/every match, and `replacement` can refer to capture groups with
`\1`, `\2`, and so on:

```{r}
.rx_replace_all("2024-01-02", "(\\d+)-(\\d+)-(\\d+)", "\\3/\\2/\\1")
```

The `.rx_remove()`/`.rx_remove_all()` functions are shorthand for
replacing with `""`:

```{r}
.rx_remove_all("order-123", "[0-9]")
```

## Splitting

The `.rx_split()` function splits each string on every match of the
pattern, keeping empty pieces rather than dropping them:

```{r}
.rx_split("a,b,,c", ",")
```

## Counting matches

The `.rx_count()` function counts non-overlapping matches per string.
This one needs a real fix, not just a wrapper: base `gregexpr()`
signals "no match" with a single `-1`, which -- read naively as a
length -- looks like *one* match rather than zero:

```{r}
.rx_count(c("banana", "kiwi", "", NA), "an")
```

## Locating matches

The `.rx_locate()`/`.rx_locate_all()` functions report the character
positions (inclusive `start`/`end`) of the first/every match:

```{r}
.rx_locate(orders, "[0-9]+")
.rx_locate_all("a1a2a3", "a")
```

## Scope: a single, PCRE-only pattern

Every function here takes exactly one pattern string, applied with
`perl = TRUE`. There's no support for stringr's `fixed()`/`coll()`/
`boundary()` pattern-type wrappers, and no locale-aware collation --
if a pattern needs to match a literal string that happens to contain
regex metacharacters, escape it yourself (e.g. with
`.rx_replace_all(x, "\\.", "_")` for a literal dot). If you need
genuinely locale-sensitive matching, this mini isn't a substitute for
`stringr` itself.
