---
title: "minimap"
---

```{r}
#| include: false
source(file.path("minimap", "minimap.R"))
```

`minimap` provides `purrr`-style mapping functions -- `mmap_map()`,
`mmap_map2()`, `mmap_imap()`, `mmap_walk()`, `mmap_iwalk()`, and the
type-stable `mmap_map_dbl()`/`mmap_map_lgl()`/`mmap_map_chr()` -- built
entirely on `lapply()`/`vapply()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minimap) for the
full API reference and scope notes; this page walks through each one,
starting from the basic case and building up to the two-input and
side-effect variants.

## Functions

### Core functions

- `mmap_map(.x, .f)` -- apply `.f` to each element of `.x`, always
  returning a list, with names preserved.
- `mmap_map_dbl()`/`mmap_map_lgl()`/`mmap_map_chr()` -- like
  `mmap_map()`, but return an atomic vector of the named type instead
  of a list, via `vapply()`.
- `mmap_map2(.x, .y, .f)` -- apply `.f` to corresponding pairs of
  elements from two same-length lists/vectors.
- `mmap_imap(.x, .f)` -- apply `.f` to each element of `.x` together
  with its name.
- `mmap_walk(.x, .f)` -- like `mmap_map()`, but for side effects only:
  returns `.x` invisibly instead of the results.
- `mmap_iwalk(.x, .f)` -- like `mmap_imap()`, but for side effects
  only: returns `.x` invisibly instead of the results.

### Internal helpers

- `.mmap_assert(expr, message)` -- a small argument-checking helper
  built on base `stop()`, used to validate inputs like matching
  lengths or the presence of names.

## `mmap_map()`

`mmap_map()` is the workhorse of the mini: apply a function to each
element of a list or atomic vector, and get back a list of results the
same length, with any names from the input preserved on the output.
It's a thin wrapper around `lapply()`, so anything you already know
about `lapply()`'s behaviour carries over directly:

```{r}
mmap_map(1:3, function(x) x + 1)
```

```{r}
mmap_map(c(a = 1, b = 2), function(x) x * 10)
```

Because the return type is always a list, `mmap_map()` is safe to use
even when `.f` returns different types or lengths for different
elements:

```{r}
mmap_map(1:3, function(x) if (x == 2) c(x, x) else x)
```

## Type-stable variants

Often you know in advance that every element of the result will be a
single number, logical, or string -- in which case getting a list back
is inconvenient. `mmap_map_dbl()`, `mmap_map_lgl()`, and
`mmap_map_chr()` cover that case: they behave like `mmap_map()`, but
return an atomic vector of the named type via `vapply()`, which also
means they *validate* that every call to `.f` really does produce a
length-1 value of the expected type:

```{r}
mmap_map_dbl(1:3, function(x) x + 1)
```

```{r}
mmap_map_lgl(1:5, function(x) x %% 2 == 0)
```

```{r}
mmap_map_chr(c(a = 1, b = 2), function(x) sprintf("value is %s", x))
```

If `.f` breaks that promise -- by returning a vector of length 2
instead of 1, say -- the call errors immediately rather than silently
producing a malformed or recycled result:

```{r}
#| error: true
mmap_map_dbl(1:3, function(x) c(x, x))
```

## `mmap_map2()` and `mmap_imap()`

Sometimes one input isn't enough. `mmap_map2()` walks two same-length
vectors in parallel, calling `.f` with one element from each:

```{r}
mmap_map2(1:3, 4:6, function(x, y) x + y)
```

Passing vectors of mismatched length is a deliberate error rather than
silent recycling:

```{r}
#| error: true
mmap_map2(1:3, 1:2, function(x, y) x + y)
```

`mmap_imap()` covers a related but different need: walking a single
named vector together with its own names, rather than a second vector.
`.f` is called with the element first, then its name:

```{r}
mmap_imap(c(a = 1, b = 2), function(val, name) paste0(name, "=", val))
```

`.x` must actually be named for `mmap_imap()` to make sense, and that's
enforced:

```{r}
#| error: true
mmap_imap(1:3, function(val, name) paste0(name, "=", val))
```

## `mmap_walk()` and `mmap_iwalk()`

`mmap_map()` and `mmap_imap()` are for when you want the *results*.
`mmap_walk()` and `mmap_iwalk()` are their side-effect-only
counterparts: same iteration behaviour, but the return value is `.x`
itself, returned invisibly, so they compose neatly at the end of a
pipeline without cluttering the console with the results of every call
to `.f`:

```{r}
mmap_walk(1:3, function(x) cat("value:", x, "\n"))
```

```{r}
mmap_iwalk(c(a = 1, b = 2), function(val, name) cat(name, "->", val, "\n"))
```

Because the input is returned invisibly, `mmap_walk()`/`mmap_iwalk()`
can be dropped into the middle of a longer expression purely for their
printing side effect, without changing what the expression evaluates
to:

```{r}
result <- mmap_walk(1:3, function(x) cat("processing", x, "\n"))
identical(result, 1:3)
```
