---
title: "minimap"
---

```{r}
#| include: false
source(file.path("minimap", "minimap.R"))
```

`minimap` provides `purrr`-style mapping functions -- `mmap_map()`,
`mmap_map2()`, `mmap_imap()`, `mmap_walk()`, `mmap_iwalk()`, and the
type-stable `mmap_map_dbl()`/`mmap_map_lgl()`/`mmap_map_chr()` -- built
entirely on `lapply()`/`vapply()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minimap) for the
full API reference and scope notes; this page walks through each in
action.

## `mmap_map()`

Applies a function to each element of a list or vector, always
returning a list, with names preserved:

```{r}
mmap_map(1:3, function(x) x + 1)
```

```{r}
mmap_map(c(a = 1, b = 2), function(x) x * 10)
```

## Type-stable variants

`mmap_map_dbl()`, `mmap_map_lgl()`, and `mmap_map_chr()` return an
atomic vector instead of a list, erroring if `.f` doesn't return a
length-1 value of the expected type for every element:

```{r}
mmap_map_dbl(1:3, function(x) x + 1)
```

```{r}
mmap_map_lgl(1:5, function(x) x %% 2 == 0)
```

```{r}
#| error: true
mmap_map_dbl(1:3, function(x) c(x, x))
```

## `mmap_map2()` and `mmap_imap()`

`mmap_map2()` walks two same-length vectors in parallel:

```{r}
mmap_map2(1:3, 4:6, function(x, y) x + y)
```

`mmap_imap()` walks a named vector together with its names, passing
`.f` the element then its name:

```{r}
mmap_imap(c(a = 1, b = 2), function(val, name) paste0(name, "=", val))
```

## `mmap_walk()` and `mmap_iwalk()`

Both run `.f` for its side effects only, returning `.x` invisibly:

```{r}
mmap_walk(1:3, function(x) cat("value:", x, "\n"))
```

```{r}
mmap_iwalk(c(a = 1, b = 2), function(val, name) cat(name, "->", val, "\n"))
```
