---
title: "minicuts"
---

```{r}
#| include: false
source(file.path("minicuts", "minicuts.R"))
```

The purpose of **minicuts** is to provide a zero-dependency
reimplementation of `cut_quantile()`/`cut_exposure_quantile()` from
[**erplots**](https://github.com/djnavarro/erplots), generalised into
`.cut_quantile()`, alongside a second function, `.cut_evenly()`,
inspired by the `chop_evenly()`/`chop_width()` pair from
[**santoku**](https://github.com/hughjonesd/santoku). Together they
cover the two common ways of cutting a numeric vector into bins: fixed
group size with data-dependent boundaries (`.cut_quantile()`), or
fixed bin geometry with a data-dependent group size
(`.cut_evenly()`). The [README](https://github.com/djnavarro/minis/tree/main/minicuts)
provides scope notes; this page walks through both functions on a
small simulated exposure variable, including tie-breaking, custom
labels, and the `exclude` argument the two functions share.

## Functions

The [**minicuts.R**](https://github.com/djnavarro/minis/blob/main/minicuts/minicuts.R)
script supplies two core functions:

- `.cut_quantile(x, n_bins = 4, exclude = NULL, exclude_label = "Excluded", ties = "upward", seed = NULL, quantile_type = 7, labeller = NULL)`
  is used to cut `x` into `n_bins` quantile bins, each holding
  (approximately) the same number of values.
- `.cut_evenly(x, n_bins = NULL, width = NULL, start = NULL, exclude = NULL, exclude_label = "Excluded", ties = "upward", labeller = NULL)`
  is used to cut `x` into equal-width bins, either a fixed number of
  them spanning `range(x)` (`n_bins`) or a fixed width with however
  many bins the data needs (`width`).

They're supported by five internal helpers, shared by both functions
-- there's no need to call any of them directly. The helpers are
these:

- `.cuts_resolve_exclude()`
- `.cuts_bin_num()`
- `.cuts_resolve_ties()`
- `.cuts_with_seed()`
- `.cuts_resolve_labels()`

## A minimal data set

A stand-in for a drug exposure variable from a clinical trial: some
patients received a placebo (exposure exactly `0`), the rest a range
of doses.

```{r}
set.seed(2984)
exposure <- c(rep(0, 12), round(rexp(48, rate = 1 / 20), 1))
head(exposure, 15)
```

## Quantile bins with `.cut_quantile()`

By default, `.cut_quantile()` cuts `x` into four bins labelled
`"Q1"`-`"Q4"`, each holding a quarter of the values:

```{r}
table(.cut_quantile(exposure))
```

Because the placebo zeroes are exact ties clustered at the low end,
they all land in `Q1` here, and the lowest break point is pulled down
to `0` as a result -- something the `exclude` argument (below) exists
to fix.

## Custom labels

The `labeller` argument accepts either a plain character vector or a
function of `n_bins` and the computed `breaks`:

```{r}
.cut_quantile(exposure, labeller = c("Low", "Mid-low", "Mid-high", "High")) |>
  table()

.cut_quantile(exposure, labeller = function(n_bins, breaks) {
  sprintf("Q%d (>%.1f)", 1:n_bins, breaks[-length(breaks)])
}) |>
  table()
```

## Ties at a quantile boundary

Repeated values sitting exactly on a quantile break point are common
with skewed or rounded data. The `ties` argument controls which bin
they land in:

```{r}
tied <- c(1, 2, 3, 4, 4, 4, 4, 5, 6, 7)
.cut_quantile(tied, n_bins = 2, ties = "upward")
.cut_quantile(tied, n_bins = 2, ties = "downward")
```

The `"split-even"` option instead divides a tied group between its two
candidate bins, aiming for equal bin sizes rather than sending every
tied value the same way. It's random, so it takes a `seed` to be
reproducible:

```{r}
.cut_quantile(tied, n_bins = 2, ties = "split-even", seed = 6017)
```

## Keeping placebo patients separate: `exclude`

Passing a predicate to `exclude` removes the matching values from the
quantile calculation entirely, so they no longer distort the break
points, but keeps them visible in the result under their own label
rather than dropping them or coding them `NA`:

```{r}
result <- .cut_quantile(exposure, exclude = function(x) x == 0, exclude_label = "Placebo")
table(result)
attr(result, "breaks")
```

The breaks are now computed purely from the 48 non-placebo doses, and
the quantile bins split that group into four roughly equal quarters,
with `"Placebo"` as a fifth, separate level.

## Equal-width bins with `.cut_evenly()`

Where `.cut_quantile()` fixes the group size and lets the bin edges
follow from the data, `.cut_evenly()` does the opposite: it fixes the
bin geometry and lets the group sizes follow from the data. Giving it
`n_bins` spans `range(x)` with that many equal-width bins:

```{r}
doses <- exposure[exposure > 0]
table(.cut_evenly(doses, n_bins = 4))
attr(.cut_evenly(doses, n_bins = 4), "breaks")
```

Giving it `width` instead fixes the bin width and works out how many
bins are needed to cover the data, anchored at `min(x)` by default:

```{r}
table(.cut_evenly(doses, width = 10))
```

## Anchoring bins with `start`

An explicit `start` moves that anchor. Here, bins are anchored at `0`
rather than `min(doses)`, so the bin edges land on round numbers:

```{r}
table(.cut_evenly(doses, width = 10, start = 0))
```

A negative `width` builds bins downward from `start` (`max(x)` by
default) instead of upward from it:

```{r}
table(.cut_evenly(doses, width = -10))
```

## Values outside the bins become `NA`

If `start` is set explicitly and doesn't reach one edge of `range(x)`,
some values fall outside every bin. `.cut_evenly()` codes those `NA`
and warns, rather than silently widening the outermost bin to cover
them:

```{r}
#| warning: true
result <- .cut_evenly(doses, width = 10, start = 5)
sum(is.na(result))
```

## `exclude` and `labeller` work the same way for both functions

The placebo patients can be kept separate here too, and custom labels
work identically to `.cut_quantile()`:

```{r}
.cut_evenly(exposure, n_bins = 4, exclude = function(x) x == 0, exclude_label = "Placebo") |>
  table()
```

## What's deliberately missing

There's no `ties = "split-even"` option for `.cut_evenly()`, and
consequently no `seed` argument either -- there's no "equal group
size" goal to chase when the bins themselves are fixed by geometry
rather than derived from data-dependent quantiles. Neither function
supports santoku's full generality: arbitrary breaks, non-numeric `x`
such as dates, or weighted quantiles. If that generality is needed,
reach for santoku itself.
