---
title: "miniverb"
---

```{r}
#| include: false
source(file.path("miniverb", "miniverb.R"))
```

The purpose of **miniverb** is to provide a zero-dependency
reimplementation of five one-table verbs from the
[**dplyr**](https://dplyr.tidyverse.org/) package: `filter()`,
`select()`, `mutate()`, `arrange()`, and `summarise()`. The
[README](https://github.com/djnavarro/minis/tree/main/miniverb)
provides full scope notes; this page walks through each verb in turn on
a small toy data frame, including a few things worth knowing about
before relying on them: how missing values are handled, what happens
when no conditions are supplied at all, and how the shared `.by`
argument groups rows for `filter()`, `mutate()`, and `summarise()`.

## Functions

The [**miniverb.R**](https://github.com/djnavarro/minis/blob/main/miniverb/miniverb.R)
script supplies:

- `.verb_filter(.data, ..., .by = NULL)` keeps rows of `.data` where
  every condition in `...` evaluates to `TRUE`. Conditions are
  evaluated in the scope of `.data`, so columns can be referred to by
  bare name.
- `.verb_select(.data, ...)` keeps (and optionally renames) columns by
  bare name, integer position, or `-` exclusion.
- `.verb_mutate(.data, ..., .by = NULL)` adds or overwrites columns via
  named `name = expr` arguments.
- `.verb_arrange(.data, ...)` reorders rows by one or more columns.
- `.verb_desc(x)` marks a column for descending order inside
  `.verb_arrange()`.
- `.verb_summarise(.data, ..., .by = NULL)` collapses to one row per
  group (or one row overall) via named `name = expr` arguments.

All five share a `.by` grouping convention (where applicable): a plain
character vector of column names, not dplyr's unquoted/tidyselect-lite
syntax.

There are also two internal helpers:

- `.verb_dotdotdot()` captures the unevaluated `...` conditions passed
  to `.verb_filter()`, so each one can be evaluated against the data
  frame individually.
- `.verb_split_by()` splits row indices into groups by `.by`'s column
  names, shared by `.verb_filter()`, `.verb_mutate()`, and
  `.verb_summarise()`.

## A minimal data set

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## `filter()`

### A single condition

Columns are referred to by bare name, evaluated in the scope of the
data frame -- there's no need to write `fruits$in_stock` or wrap the
call in `with()`, as you would for [**minicase**](../minicase/)'s
`.case_when()`:

```{r}
.verb_filter(fruits, in_stock)
```

### Multiple conditions combine with AND

Passing more than one condition is equivalent to combining them with
`&`, so a row is kept only if every condition holds. This is often more
readable than a single expression joined with `&&`/`&`, especially as
the number of conditions grows:

```{r}
.verb_filter(fruits, in_stock, price < 4)
```

That's equivalent to writing the conditions out explicitly and
combining them yourself:

```{r}
identical(
  .verb_filter(fruits, in_stock, price < 4),
  fruits[fruits$in_stock & fruits$price < 4 & !is.na(fruits$price), ]
)
```

### `NA` conditions drop the row

The `date` row has a missing `price`, so any condition involving
`price` is `NA` for that row rather than `TRUE`/`FALSE`. The
`.verb_filter()` function treats `NA` the same way `dplyr::filter()`
does: as "not proven to be kept", so the row is dropped rather than
kept by default or causing an error:

```{r}
.verb_filter(fruits, price > 1)
```

Notice that `date` disappears from the result entirely, rather than
appearing with `NA` values or being kept because "we don't know" --
this is a common surprise for anyone used to base R's `[` subsetting,
where an `NA` index instead produces a row of `NA`s:

```{r}
fruits[fruits$price > 1, ]
```

### No conditions returns the data frame unchanged

Calling `.verb_filter()` with no conditions at all is legitimate and
simply leaves the data frame unchanged -- useful when conditions are
built up programmatically and might end up empty:

```{r}
identical(.verb_filter(fruits), fruits)
```

### Grouped conditions with `.by`

The `.by` argument takes a character vector of column names and
evaluates the conditions separately within each group -- useful when a
condition depends on a per-group quantity like a group mean:

```{r}
.verb_filter(fruits, price > mean(price, na.rm = TRUE), .by = "in_stock")
```

## `select()`

### Keeping and renaming columns

`.verb_select()` keeps columns in the order they're listed, and a
`new = old` argument renames a column on the way through:

```{r}
.verb_select(fruits, name, cost = price)
```

### Dropping columns instead

A `-` in front of a column name drops it rather than keeping it, which
reads more naturally than listing every column to keep when only one
or two need to go. Positive and negative selections can't be mixed in
the same call:

```{r}
.verb_select(fruits, -in_stock)
```

## `mutate()`

### Adding a new column

`.verb_mutate()` takes named `name = expr` arguments and adds each one
as a new column, evaluated in the scope of `fruits`:

```{r}
.verb_mutate(fruits, discounted = round(price * 0.9, 2))
```

### Later columns can use earlier ones

Arguments are evaluated left to right, so an expression can refer to a
column added by an earlier argument in the same call. Below, `savings`
uses `discounted` even though it doesn't exist in `fruits` itself:

```{r}
.verb_mutate(
  fruits,
  discounted = round(price * 0.9, 2),
  savings = round(price - discounted, 2)
)
```

### Grouped columns with `.by`

Passing `.by` evaluates each expression separately within each group,
then puts the results back in the original row order -- here, each
fruit's price is compared to the average price within its own
`in_stock` group rather than the average across all five fruits:

```{r}
.verb_mutate(
  fruits,
  price_vs_group = round(price - mean(price, na.rm = TRUE), 2),
  .by = "in_stock"
)
```

## `arrange()`

### Ascending by default, descending with `.verb_desc()`

`.verb_arrange()` sorts ascending by default. Wrapping a column in
`.verb_desc()` sorts it descending instead, and `NA`s sort last either
way, matching `dplyr::arrange()`:

```{r}
.verb_arrange(fruits, .verb_desc(price))
```

### Breaking ties with a second column

Extra arguments break ties left over from earlier ones, the same way
`order()` does. Sorting by `in_stock` first and price second groups the
out-of-stock `cherry` row on its own, then orders the remaining rows by
descending price:

```{r}
.verb_arrange(fruits, in_stock, .verb_desc(price))
```

## `summarise()`

### One row overall

With no `.by`, `.verb_summarise()` collapses `fruits` down to a single
row, computing each named expression once over every row:

```{r}
.verb_summarise(fruits, average_price = round(mean(price, na.rm = TRUE), 2), n = length(name))
```

### One row per group with `.by`

With `.by`, one summary row is produced per group instead, and the
grouping columns appear first in the result:

```{r}
.verb_summarise(
  fruits,
  average_price = round(mean(price, na.rm = TRUE), 2),
  n = length(name),
  .by = "in_stock"
)
```
