---
title: "minicli"
---

```{r}
#| include: false
source(file.path("minicli", "minicli.R"))
```

`minicli` is a minimal stand-in for the small slice of `cli` most
packages actually use: coloured alert messages, unicode symbols with
an ascii fallback, a rule, and a bullet list. See the
[README](https://github.com/djnavarro/minis/tree/main/minicli) for the
full API reference and capability-detection details; this page walks
through each piece.

One thing to note about the output below: Quarto/knitr renders are
detected as a non-terminal (`getOption("knitr.in.progress")`), so ANSI
colour is automatically disabled here -- the alerts below print in
plain text with their symbol prefix, exactly as they would in a
captured log or a non-interactive session, rather than showing colour
codes as literal text.

## Alerts

`mcli_alert_success()`/`_danger()`/`_warning()`/`_info()` are thin
`message()` wrappers with a symbol prefix; extra `...` arguments are
passed through `sprintf()`:

```{r}
mcli_alert_success("Wrote %d files to %s", 3, "output/")
```

```{r}
mcli_alert_warning("Skipped %d rows with missing values", 12)
```

```{r}
mcli_alert_danger("Failed to connect to %s", "database")
```

```{r}
mcli_alert_info("Using default configuration")
```

## Symbols

`mcli_symbol()` looks up a named symbol, respecting unicode
availability:

```{r}
mcli_symbol("tick")
mcli_symbol("cross")
mcli_symbol("arrow_right")
```

## Rule

`mcli_rule()` prints a horizontal divider, optionally with a centred
title:

```{r}
mcli_rule("Summary")
```

## Bullets

`mcli_bullets()` prints a simple bulleted list, one entry per line:

```{r}
mcli_bullets(c("one", "two", "three"))
```

## Colours and styles

`mcli_col_*()`/`mcli_style_*()` wrap text in an ANSI code, and are
no-ops (returning the text unchanged) whenever ANSI is disabled --
which, per the note above, is always true on this page:

```{r}
identical(mcli_col_red("danger"), "danger")
identical(mcli_style_bold("emphasis"), "emphasis")
```
