---
title: "minicli"
---

```{r}
#| include: false
source(file.path("minicli", "minicli.R"))
```

`minicli` is a minimal stand-in for the small slice of `cli` most
packages actually use: coloured alert messages, unicode symbols with
an ascii fallback, a rule, and a bullet list. See the
[README](https://github.com/djnavarro/minis/tree/main/minicli) for the
full API reference and capability-detection details; this page walks
through each piece and builds up to how they combine.

## Functions

### Core functions

- `mcli_alert_success()`/`_danger()`/`_warning()`/`_info()` -- print a
  `message()` with a coloured symbol prefix; extra `...` arguments are
  passed through `sprintf()`.
- `mcli_symbol(name)` -- look up a single symbol by name, returning its
  unicode glyph or an ascii fallback depending on what's safe to print.
- `mcli_rule(title = NULL)` -- print a horizontal divider spanning the
  console width, optionally with a centred title.
- `mcli_bullets(items)` -- print a simple bulleted list, one entry per
  line.
- `mcli_col_black()`/`_red()`/`_green()`/`_yellow()`/`_blue()`/
  `_magenta()`/`_cyan()`/`_white()`/`_grey()` -- wrap a string in an
  ANSI colour code.
- `mcli_style_bold()`/`_dim()`/`_italic()`/`_underline()` -- wrap a
  string in an ANSI style code.

### Internal helpers

- `.mcli_ansi_enabled()` -- decide whether it's currently safe to emit
  ANSI escape codes (checks `cli.num_colors`, `NO_COLOR`, whether
  knitr/Quarto is rendering, whether output is being captured, whether
  it's running in Positron's or RStudio's Console pane, and whether
  stdout is a real terminal).
- `.mcli_positron_console_with_color()`/
  `.mcli_rstudio_console_with_color()` -- recognise Positron's and
  RStudio's Console panes explicitly, since neither is a real tty but
  both do support ANSI colour.
- `.mcli_unicode_enabled()` -- decide whether it's currently safe to
  print unicode symbols (checks `cli.unicode`, then the session's
  locale).
- `.mcli_ansi_style(code)` -- factory that builds one of the
  `mcli_col_*()`/`mcli_style_*()` functions from a raw SGR code.
- `.mcli_alert(symbol_name, color_fn, text, ...)` -- shared
  implementation behind the four `mcli_alert_*()` functions.

## Capability detection

Every colour/symbol decision in `minicli` runs through
`.mcli_ansi_enabled()` and `.mcli_unicode_enabled()`, so the same code
looks right whether it's running in an interactive terminal that
supports unicode and colour, Positron's or RStudio's Console pane
(neither of which is a real tty, but both of which do support ANSI
colour), a plain terminal that supports neither, or -- as on this page
-- a Quarto render. Quarto/knitr renders are detected as a non-terminal
(`getOption("knitr.in.progress")`), so ANSI colour is disabled
automatically here. That means the alerts below print in plain text
with their symbol prefix, exactly as they would in a captured log or a
non-interactive session, rather than showing colour codes as literal
text on the page.

## Alerts

`mcli_alert_success()`/`_danger()`/`_warning()`/`_info()` are the
functions you'll reach for most often: thin `message()` wrappers with a
symbol prefix, meant for the kind of status line a script prints while
it works. Extra `...` arguments are passed through `sprintf()`, so the
first argument can be a format string:

```{r}
mcli_alert_success("Wrote %d files to %s", 3, "output/")
```

```{r}
mcli_alert_warning("Skipped %d rows with missing values", 12)
```

```{r}
mcli_alert_danger("Failed to connect to %s", "database")
```

```{r}
mcli_alert_info("Using default configuration")
```

If there's no `...` to format with, the text is used as-is -- no
`sprintf()` call, so a literal `%` in the message is safe:

```{r}
mcli_alert_info("Discount: 10% off")
```

## Symbols

Each alert's prefix comes from `mcli_symbol()`, which looks up a named
symbol and returns either its unicode glyph or an ascii fallback,
depending on `.mcli_unicode_enabled()`. It's also useful on its own,
for building custom messages that follow the same convention as the
built-in alerts:

```{r}
mcli_symbol("tick")
mcli_symbol("cross")
mcli_symbol("arrow_right")
```

An unrecognised name is an error rather than a silent `NA`:

```{r}
#| error: true
mcli_symbol("star")
```

## Rule

`mcli_rule()` prints a horizontal divider, useful for separating
sections of console output. Called with no arguments, it spans the
full console width:

```{r}
mcli_rule()
```

Pass a `title` and it's centred within the rule instead:

```{r}
mcli_rule("Summary")
```

## Bullets

`mcli_bullets()` prints a simple bulleted list, one entry per line --
handy for summarising a small set of results without reaching for a
full table:

```{r}
mcli_bullets(c("one", "two", "three"))
```

## Colours and styles

Under the hood, each alert calls one of the `mcli_col_*()` functions to
colour its symbol. Those functions -- along with the `mcli_style_*()`
family -- are also exported directly, for colouring or styling
arbitrary text of your own. Both wrap their input in an ANSI escape
code and its matching reset code, and both are no-ops (returning the
text unchanged) whenever `.mcli_ansi_enabled()` is `FALSE` -- which, per
the capability-detection note above, is always true on this page:

```{r}
identical(mcli_col_red("danger"), "danger")
identical(mcli_style_bold("emphasis"), "emphasis")
```

Colours and styles nest freely, since each wrapper's reset code is
emitted immediately after its own content rather than at the very end
of the whole string:

```{r}
mcli_style_bold(mcli_col_red("bold and red"))
```

Outside of a Quarto render -- in an interactive terminal session, say
-- that same call would print bold red text rather than the plain
string shown here.
