---
title: "minicase"
---

```{r}
#| include: false
source(file.path("minicase", "minicase.R"))
```

`minicase` provides a single function, `mcase_case_when()`, a minimal
reimplementation of `dplyr::case_when()`: vectorised if/else across a
sequence of `condition ~ value` formulas, first match wins. See the
[README](https://github.com/djnavarro/minis/tree/main/minicase) for
scope notes; this page walks through it on the same toy fruits data
frame used in the [minifilter vignette](../minifilter/).

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## No data-masking: reference columns with `$`, or wrap in `with()`

Unlike `dplyr::case_when()`, conditions and values are evaluated in
the *caller's* environment, not the data frame's -- so bare column
names don't resolve on their own:

```{r}
with(fruits, mcase_case_when(
  price < 1 ~ "cheap",
  price < 3 ~ "mid",
  price >= 3 ~ "pricey"
))
```

## First match wins

`cherry` (price 3.00) fails `price < 3` but matches `price >= 3`;
formulas are checked in order, and the first `TRUE` condition for each
position determines its value even if a later condition would also
match.

## Unmatched positions are `NA`

`date` has a missing price, so every condition above evaluates to `NA`
for that row, and `date` itself gets `NA` in the result -- there's no
condition it definitively satisfies. Add a `TRUE ~ ...` formula at the
end as a catch-all default, the same way you would in `dplyr`:

```{r}
with(fruits, mcase_case_when(
  price < 1  ~ "cheap",
  price < 3  ~ "mid",
  price >= 3 ~ "pricey",
  TRUE       ~ "unknown"
))
```

## Scope: no `.default`/`.ptype`/`.size`

That trailing `TRUE ~ "unknown"` formula *is* the idiomatic way to get
a default in both `dplyr::case_when()` and `mcase_case_when()` --
`minicase` just doesn't implement dplyr's newer, separate `.default`
argument as an alternative spelling of the same thing.
