---
title: "minicase"
---

```{r}
#| include: false
source(file.path("minicase", "minicase.R"))
```

The purpose of **minicase** is to provide a zero-dependency
reimplementation of a single function from the
[**dplyr**](https://dplyr.tidyverse.org/) package: `case_when()`. It
performs a vectorised if/else across a sequence of `condition ~ value`
formulas, where the first match wins. The [README](https://github.com/djnavarro/minis/tree/main/minicase)
provides scope notes; this page walks through it on the same toy
fruits data frame used in the [**miniverb**](../miniverb/)
vignette, building up to how missing values and default values
interact.

## Functions

The [**minicase.R**](https://github.com/djnavarro/minis/blob/main/minicase/minicase.R)
script supplies one core function:

- `.case_when(...)` is used to take a sequence of `condition ~ value`
  formulas and return a vector the same length as the (recycled)
  conditions/values, taking the value from the first formula whose
  condition is `TRUE` at each position, and `NA` wherever no condition
  matched.

It is supported by five internal helpers, all of which support
`.case_when()`'s recycling and type-consistency checks -- there's no
need to call any of them directly. The helpers are these:

- `.case_validate_length()`
- `.case_replace_with()`
- `.case_check_length()`
- `.case_check_type()`
- `.case_check_class()`

## A minimal data set

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## No data-masking

There is no data-masking in this mini: you must reference columns with `$`, or wrap in `with()`. Unlike `dplyr::case_when()`, conditions and values are evaluated in the *caller's* environment, not the data frame's -- so bare column names don't resolve on their own. The idiomatic way around that, mirrored here from base R, is to wrap the whole call in `with(data, ...)`, which temporarily makes the data frame's columns available as bare names for the duration of the call:

```{r}
with(fruits, .case_when(
  price < 1 ~ "cheap",
  price < 3 ~ "mid",
  price >= 3 ~ "pricey"
))
```

Referring to columns via `fruits$price` directly, without `with()`, works exactly the same way -- `with()` is a convenience, not a requirement:

```{r}
identical(
  with(fruits, .case_when(price < 3 ~ "cheap", TRUE ~ "not cheap")),
  .case_when(fruits$price < 3 ~ "cheap", TRUE ~ "not cheap")
)
```

## First match wins

Formulas are checked in order, and the first `TRUE` condition for each position determines its value, even if a later condition would also match. The `cherry` row (price 3.00) is a good illustration: it fails
`price < 3`, but matches `price >= 3`, so it's classified as
`"pricey"` above. Conditions can combine several tests with `&`, which is useful for layering more specific rules ahead of more general ones:

```{r}
with(fruits, .case_when(
  in_stock & price < 1 ~ "cheap and in stock",
  in_stock             ~ "in stock",
  TRUE                 ~ "out of stock or unknown price"
))
```

The `banana` row matches both the first and second formula here (it's
in stock *and* under $1), but because the first-match-wins rule checks
formulas top to bottom, it's classified by the more specific
`"cheap and in stock"` rule rather than the more general `"in stock"`
one. Ordering formulas from most to least specific is exactly how you
get that behaviour deliberately, rather than by accident.

## Unmatched positions are `NA`

The `date` row has a missing price, so every condition involving
`price` evaluates to `NA` for that row -- neither `TRUE` nor `FALSE`
-- and `date` itself gets `NA` in the result, since there's no
condition it definitively satisfies:

```{r}
with(fruits, .case_when(
  price < 1 ~ "cheap",
  price < 3 ~ "mid",
  price >= 3 ~ "pricey"
))
```

Add a `TRUE ~ ...` formula at the end as a catch-all default, the same way you would in `dplyr`, and every remaining position -- including ones where every prior condition was `NA` rather than `FALSE` -- picks up that value instead:

```{r}
with(fruits, .case_when(
  price < 1  ~ "cheap",
  price < 3  ~ "mid",
  price >= 3 ~ "pricey",
  TRUE       ~ "unknown"
))
```

## Factors must share the same levels, not just the same class

Values across formulas are also checked for type/class consistency.
For factors specifically, sharing class `"factor"` isn't enough --
the levels must match too, since assigning a factor label that isn't
among an existing factor's levels would otherwise silently produce
`NA` rather than raising a visible error:

```{r}
#| error: true
f1 <- factor(c("a", "b"), levels = c("a", "b"))
f2 <- factor(c("x", "y"), levels = c("x", "y"))
.case_when(c(TRUE, FALSE) ~ f1, c(FALSE, TRUE) ~ f2)
```

Two factors with the same levels (in the same order) combine without
any trouble:

```{r}
f3 <- factor(c("a", "b"), levels = c("a", "b"))
f4 <- factor(c("b", "a"), levels = c("a", "b"))
.case_when(c(TRUE, FALSE) ~ f3, c(FALSE, TRUE) ~ f4)
```

## Scope: no `.default`/`.ptype`/`.size`

The mini does not implement any of these arguments. The trailing `TRUE ~ "unknown"` formula in the previous example *is* the idiomatic way to get a default value, and is in fact very often used when writing code with the real `dplyr::case_when()` function. There was no need for this mini to implement dplyr's newer, separate `.default` argument as an alternative spelling of the same thing, and little to be gained by adding `.ptype`/`.size` to control the output type/length explicitly. If your use case needs those, a plain `TRUE ~ value` formula covers `.default`, and the output length/type simply follow from the conditions. If that still isn't sufficient, then it's likely you need the real version, not the mini.
