---
title: "minicase"
---

```{r}
#| include: false
source(file.path("minicase", "minicase.R"))
```

`minicase` provides a single function, `mcase_case_when()`, a minimal reimplementation of `dplyr::case_when()`: vectorised if/else across a sequence of `condition ~ value` formulas, first match wins. See the [README](https://github.com/djnavarro/minis/tree/main/minicase) for scope notes; this page walks through it on the same toy fruits data frame used in the [minifilter vignette](../minifilter/), building up to how missing values and default values interact.

## Functions

There is one core function:

- `mcase_case_when(...)` -- given a sequence of `condition ~ value`
  formulas, return a vector the same length as the (recycled)
  conditions/values, taking the value from the first formula whose
  condition is `TRUE` at each position, and `NA` wherever no condition
  matched.

It is supported by five internal helpers, all supporting `mcase_case_when()`'s recycling and type-consistency checks -- there's no need to call any of them directly. The helpers are these:

- `.mcase_validate_length()`
- `.mcase_replace_with()`
- `.mcase_check_length()`
- `.mcase_check_type()`
- `.mcase_check_class()`

## A minimal data set

```{r}
fruits <- data.frame(
  name     = c("apple", "banana", "cherry", "date", "elderberry"),
  price    = c(1.20, 0.50, 3.00, NA, 4.50),
  in_stock = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)
fruits
```

## No data-masking

There is no data-masking in this mini: you must reference columns with `$`, or wrap in `with()`. Unlike `dplyr::case_when()`, conditions and values are evaluated in the *caller's* environment, not the data frame's -- so bare column names don't resolve on their own. The idiomatic way around that mirrored here from base R, is to wrap the whole call in `with(data, ...)` which temporarily makes the data frame's columns available as bare names for the duration of the call:

```{r}
with(fruits, mcase_case_when(
  price < 1 ~ "cheap",
  price < 3 ~ "mid",
  price >= 3 ~ "pricey"
))
```

Referring to columns via `fruits$price` directly, without `with()`, works exactly the same way -- `with()` is a convenience, not a requirement:

```{r}
identical(
  with(fruits, mcase_case_when(price < 3 ~ "cheap", TRUE ~ "not cheap")),
  mcase_case_when(fruits$price < 3 ~ "cheap", TRUE ~ "not cheap")
)
```

## First match wins

Formulas are checked in order, and the first `TRUE` condition for each position determines its value, even if a later condition would also match. `cherry` (price 3.00) is a good illustration: it fails `price < 3`, but matches `price >= 3`, so it's classified as `"pricey"` above. Conditions can combine several tests with `&`, which is useful for layering more specific rules ahead of more general ones:

```{r}
with(fruits, mcase_case_when(
  in_stock & price < 1 ~ "cheap and in stock",
  in_stock             ~ "in stock",
  TRUE                 ~ "out of stock or unknown price"
))
```

`banana` matches both the first and second formula here (it's in stock *and* under $1), but because the first-match-wins rule checks formulas top to bottom, it's classified by the more specific `"cheap and in
stock"` rule rather than the more general `"in stock"` one. Ordering formulas from most to least specific is exactly how you get that behaviour deliberately, rather than by accident.

## Unmatched positions are `NA`

`date` has a missing price, so every condition involving `price` evaluates to `NA` for that row -- neither `TRUE` nor `FALSE` -- and `date` itself gets `NA` in the result, since there's no condition it definitively satisfies:

```{r}
with(fruits, mcase_case_when(
  price < 1 ~ "cheap",
  price < 3 ~ "mid",
  price >= 3 ~ "pricey"
))
```

Add a `TRUE ~ ...` formula at the end as a catch-all default, the same way you would in `dplyr`, and every remaining position -- including ones where every prior condition was `NA` rather than `FALSE` -- picks up that value instead:

```{r}
with(fruits, mcase_case_when(
  price < 1  ~ "cheap",
  price < 3  ~ "mid",
  price >= 3 ~ "pricey",
  TRUE       ~ "unknown"
))
```

## Scope: no `.default`/`.ptype`/`.size`

The mini does not implement any of these arguments. The trailing `TRUE ~ "unknown"` formula in the previous example *is* the idiomatic way to get a default value (and indeed is very often used in when writing code with the real `dplyr::case_when()` function). I simply decided there was no need for this mini to implement dplyr's newer, separate `.default` argument as an alternative spelling of the same thing, nor did I think there was much to be gained by adding `.ptype`/ `.size` to control the output type/length explicitly. If your use case needs those, a plain `TRUE ~ value` formula covers `.default`, and the output length/type simply follow from whatever the conditions. If that still isn't sufficient, then it's likely you need the real version not the mini. 
