---
title: "ministr"
---

```{r}
#| include: false
source(file.path("ministr", "ministr.R"))
```

The purpose of **ministr** is to provide a zero-dependency
reimplementation of stringr's basic string manipulation verbs from the
[**stringr**](https://stringr.tidyverse.org/) package: padding,
trimming, whitespace squishing, substrings, length, case conversion,
duplication, concatenation, and wrapping. The [README](https://github.com/djnavarro/minis/tree/main/ministr)
provides scope notes; this page walks through each function on a small
set of toy strings, including two behaviours worth knowing about
before you rely on it: how `NA` is propagated through concatenation,
and how title case is found without a Unicode-aware word-break
algorithm. (Pattern-matching functions like `str_detect()`/
`str_replace()` are out of scope here -- they belong to a separate,
companion mini.)

## Functions

The [**ministr.R**](https://github.com/djnavarro/minis/blob/main/ministr/ministr.R)
script supplies eleven core functions:

- `.str_pad(string, width, side, pad)` is used to pad a string to a
  fixed width.
- `.str_trim(string, side)` is used to trim leading/trailing
  whitespace.
- `.str_squish(string)` is used to trim and collapse internal
  whitespace runs to a single space.
- `.str_sub(string, start, end)` is used to extract a substring by
  (possibly negative) character position.
- `.str_length(string)` is used to count the characters in a string.
- `.str_to_upper(string)` and `.str_to_lower(string)` convert case.
- `.str_to_title(string)` converts to title case.
- `.str_to_sentence(string)` converts to sentence case.
- `.str_dup(string, times)` is used to repeat a string.
- `.str_c(..., sep, collapse)` is used to concatenate strings,
  propagating `NA`.
- `.str_wrap(string, width, indent, exdent)` is used to wrap a string
  to a target line width.

It contains no internal helpers.

## Padding

The `.str_pad()` function pads to a fixed width, left by default:

```{r}
.str_pad(c("7", "70", "700"), width = 4, pad = "0")
```

Padding can go on the right, or on both sides at once (with any leftover
padding placed on the right when the total isn't evenly split):

```{r}
.str_pad("hi", width = 6, side = "right", pad = "-")
.str_pad("hi", width = 7, side = "both", pad = "-")
```

A string already at least as wide as `width` is returned unchanged, and
an `NA` input stays `NA` rather than being padded. The rest of this
page reuses this small toy vector, mixing untrimmed and inconsistently
cased strings with a missing value:

```{r}
names <- c("  bilbo baggins", "FRODO", NA)
.str_pad(names, width = 20)
```

## Trimming and squishing whitespace

The `.str_trim()` function trims both sides by default, or a single
side on request:

```{r}
.str_trim(names[1])
.str_trim(names[1], side = "left")
```

The `.str_squish()` function goes further, also collapsing any run of
internal whitespace down to a single space:

```{r}
.str_squish("  too   much   space  ")
```

## Substrings

The `.str_sub()` function extracts by character position, and accepts
negative positions counting back from the end of the string, the same
way `stringr::str_sub()` does:

```{r}
.str_sub("hello world", 1, 5)
.str_sub("hello world", -5, -1)
```

Positions past either end of the string are clamped rather than
erroring, and a start after the end just gives an empty string:

```{r}
.str_sub("hi", 1, 10)
.str_sub("hello", 4, 2)
```

## Length

The `.str_length()` function counts characters, and an `NA` input
gives an `NA` count rather than treating it as the two-character
string `"NA"`:

```{r}
.str_length(names)
```

## Case conversion

The `.str_to_upper()`/`.str_to_lower()` functions are plain
upper/lower casing:

```{r}
.str_to_upper("Bilbo")
.str_to_lower("Bilbo")
```

The `.str_to_title()` function lower-cases the whole string first, then
upper-cases the first letter of each word:

```{r}
.str_to_title("  bilbo baggins")
```

The `.str_to_sentence()` function is similar, but only upper-cases the
very first letter of the string -- the rest is lower-cased regardless
of word boundaries, since the whole input is treated as a single
sentence rather than split on `.`/`!`/`?`:

```{r}
.str_to_sentence("BILBO BAGGINS the hobbit")
```

Word starts for `.str_to_title()` are found with the ASCII regex `\\b`,
not a Unicode-aware word-break algorithm. That's a fine approximation
for whitespace-delimited Latin text, but it isn't a substitute for
stringi's locale-aware title casing if your strings involve other
scripts.

## Duplication

The `.str_dup()` function repeats each string, recycling `times`
against `string`:

```{r}
.str_dup("ab", 3)
.str_dup(c("x", "y"), c(2, 4))
```

## Concatenation and `NA` propagation

The `.str_c()` function behaves like `paste0()` with a separator, but
an `NA` anywhere in the inputs to a given position makes that
position's result `NA`:

```{r}
.str_c("Mr. ", names, sep = "")
```

Compare that with what base `paste0()` does with the same input --
it coerces the missing name to the literal string `"NA"` instead of
propagating the missing value:

```{r}
paste0("Mr. ", names)
```

Supplying `collapse` joins everything into one string, and an `NA`
anywhere in the collapsed inputs makes the whole collapsed result
`NA`:

```{r}
.str_c(c("bilbo", "frodo"), collapse = ", ")
.str_c(c("bilbo", NA), collapse = ", ")
```

## Wrapping

The `.str_wrap()` function rewraps a string to a target line width,
returning a single string per input with lines joined by `"\n"`:

```{r}
wrapped <- .str_wrap(
  "Bilbo Baggins was a very well-to-do hobbit who lived in a hole in the ground.",
  width = 20
)
cat(wrapped)
```

## Scope: no locale awareness

None of these functions are locale-aware, unlike their `stringr`
counterparts (which delegate to stringi/ICU under the hood). That's
a deliberate trade-off: it keeps this mini free of a regex-engine or
ICU dependency, at the cost of correctness on non-Latin scripts and
locale-specific casing rules (Turkish dotless *i*, for example). If
your strings need that, this mini isn't a substitute for `stringr`
itself.
