---
title: "minipivot"
---

```{r}
#| include: false
source(file.path("minipivot", "minipivot.R"))
```

`minipivot` provides two functions, `.pivot_longer()` and
`.pivot_wider()`, minimal reimplementations of `tidyr::pivot_longer()`
and `tidyr::pivot_wider()`: reshaping a data frame between "long" and
"wide" layouts. See the
[README](https://github.com/djnavarro/minis/tree/main/minipivot) for
scope notes; this page walks through both functions on a small toy
data frame, including the column-selection styles `.pivot_longer()`
supports and what happens when `.pivot_wider()` is asked to spread
duplicate rows into one cell.

## Functions

- `.pivot_longer(.data, cols, names_to = "name", values_to = "value")`
  -- melt `cols` into two new columns holding the pivoted column names
  and their values.
- `.pivot_wider(.data, names_from, values_from, values_fill = NULL)`
  -- spread the unique values of `names_from` into new columns, filled
  from `values_from`.

## A minimal data set

```{r}
fish <- data.frame(
  location = c("lake", "lake", "sea", "sea"),
  species  = c("trout", "bass", "trout", "bass"),
  count    = c(5, 3, 7, 2)
)
fish
```

## Wide to long

`.pivot_longer()` needs to know which columns to pivot. Here, `count`
is the only column with values to melt, and `location`/`species`
along for the ride as id columns:

```{r}
.pivot_longer(fish, count, names_to = "metric", values_to = "n")
```

## Long to wide

`.pivot_wider()` spreads `species`'s unique values (`trout`/`bass`)
into new columns, filled from `count`. There's no `id_cols` argument
to set -- every column other than `names_from`/`values_from`
(`location`, here) is treated as an id column automatically:

```{r}
wide <- .pivot_wider(fish, names_from = "species", values_from = "count")
wide
```

## Round-tripping

Pivoting the wide result back out reproduces the original long data
(row order aside, since `.pivot_wider()` reorders rows by id column):

```{r}
.pivot_longer(wide, c(trout, bass), names_to = "species", values_to = "count")
```

## Selecting `cols` a few different ways

`.pivot_longer()`'s `cols` accepts more than just a single bare name.
Bare names combined with `c(...)`, a plain character vector, and a
negated `-c(...)` ("every column except these") all agree, since
`trout`/`bass` are the only two columns not already used as an id
column above:

```{r}
identical(
  .pivot_longer(wide, c(trout, bass)),
  .pivot_longer(wide, c("trout", "bass"))
)
identical(
  .pivot_longer(wide, c(trout, bass)),
  .pivot_longer(wide, -location)
)
```

There is no tidyselect-style `starts_with()`/`everything()` support --
just bare names, `c(...)`, character/numeric vectors, and unary `-`
negation.

## Missing combinations: `values_fill`

Dropping a row before widening leaves a combination unrepresented --
`sea`/`bass` is missing below, so it comes back as `NA` by default:

```{r}
.pivot_wider(fish[-4, ], names_from = "species", values_from = "count")
```

`values_fill` supplies a value to use instead:

```{r}
.pivot_wider(fish[-4, ], names_from = "species", values_from = "count", values_fill = 0)
```

## Duplicate combinations are an error, not a list-column

Real `tidyr::pivot_wider()` handles a duplicate `names_from`/id
combination by warning and packing the extra values into a
list-column. `minipivot` has no list-column support (matching
[`minitable`](../minitable/)'s scope), so the same situation is a hard
error instead:

```{r}
#| error: true
dup <- rbind(fish, data.frame(location = "lake", species = "trout", count = 99))
.pivot_wider(dup, names_from = "species", values_from = "count")
```
