---
title: "minitrap"
---

```{r}
#| include: false
source(file.path("minitrap", "minitrap.R"))
```

`minitrap` provides two `purrr`-style function adverbs, `mtrap_safely()`
and `mtrap_quietly()`, that trap errors or console output instead of
letting them reach the console. See the
[README](https://github.com/djnavarro/minis/tree/main/minitrap) for the
full API reference and scope notes; this page walks through both in
action.

## `mtrap_safely()`

Wrap a function so a call that would normally error instead returns a
list with the error captured alongside a `NULL` result:

```{r}
safe_log <- mtrap_safely(log)
safe_log(-1)
```

A successful call returns the result with `error` set to `NULL`:

```{r}
safe_log(10)
```

## `mtrap_quietly()`

Wrap a function so printed output, warnings, and messages are captured
instead of shown:

```{r}
quiet_fn <- mtrap_quietly(function() {
  message("starting")
  warning("using a default")
  42
})
quiet_fn()
```

`mtrap_quietly()` does not catch errors -- compose with
`mtrap_safely()` for that:

```{r}
robust_fn <- mtrap_safely(mtrap_quietly(function() stop("boom")))
robust_fn()
```
