---
title: "minitrap"
---

```{r}
#| include: false
source(file.path("minitrap", "minitrap.R"))
```

`minitrap` provides two `purrr`-style function adverbs, `.trap_safely()`
and `.trap_quietly()`. A function adverb takes a function and returns a
*new* function with modified behaviour -- here, one that traps errors
or console output instead of letting them propagate or print. See the
[README](https://github.com/djnavarro/minis/tree/main/minitrap) for the
full API reference and scope notes; this page walks through both in
action, and how they compose together.

## Functions

- `.trap_safely(.f)` -- wrap `.f` so that a call which would normally
  throw an error instead returns `list(result = ..., error = ...)`,
  with exactly one of the two `NULL`.
- `.trap_quietly(.f)` -- wrap `.f` so that printed output, warnings,
  and messages are captured instead of shown, returning
  `list(result = ..., output = ..., warnings = ..., messages = ...)`.

`minitrap` has no internal helpers -- both functions are short enough
to be self-contained.

## `.trap_safely()`

Ordinarily, a function that errors stops execution and requires you to
wrap every call site in `tryCatch()` if you want to keep going.
`.trap_safely()` inverts that: it takes a function and returns a new
one that never throws, instead capturing the error as data. Calling
`.trap_safely(log)` on a non-numeric input, for example, doesn't let
the resulting error reach the console -- it comes back as a list with
`result = NULL` and `error` set to the condition object that would have
been signalled:

```{r}
safe_log <- .trap_safely(log)
safe_log("a")
```

A successful call returns the result with `error` set to `NULL`
instead, so the two fields are always mutually exclusive -- checking
`is.null(out$error)` tells you unambiguously whether the call
succeeded:

```{r}
safe_log(10)
```

This is particularly useful when mapping a function over many inputs
and some of them are expected to fail -- rather than the whole
iteration aborting on the first error, each element's outcome (success
or failure) is captured individually. Combining `.trap_safely()` with
[`minimap`](../minimap/)'s `.iter_map()` gives exactly that pattern
(this page sources `minimap.R` too, purely to demonstrate the two
minis working together -- `minitrap` itself has no dependency on
`minimap`):

```{r}
#| include: false
source(file.path("minimap", "minimap.R"))
```

```{r}
results <- .iter_map(list(4, "a", 9), .trap_safely(sqrt))
.iter_map_lgl(results, function(r) is.null(r$error))
```

## `.trap_quietly()`

Where `.trap_safely()` is about errors, `.trap_quietly()` is about
everything a function might print along the way: `cat()`/`print()`
output, `warning()`s, and `message()`s. Rather than letting any of that
reach the console, `.trap_quietly()` collects it all and hands it back
as part of the return value:

```{r}
quiet_fn <- .trap_quietly(function() {
  message("starting")
  warning("using a default")
  42
})
quiet_fn()
```

Note the shape of the result: `result` holds the ordinary return value
(`42`), `warnings` and `messages` are character vectors with one entry
per condition raised (in the order they occurred), and `output` is
whatever was printed via `cat()`/`print()`, collapsed into a single
string -- empty here, since this function didn't print anything
directly:

```{r}
loud_fn <- .trap_quietly(function() {
  cat("computing...\n")
  99
})
loud_fn()
```

Crucially, `.trap_quietly()` does *not* catch errors -- a function that
errors still aborts the call, same as if it were called directly.
Compose it with `.trap_safely()` when you want both output-trapping and
error-trapping at once:

```{r}
#| error: true
.trap_quietly(function() stop("boom"))()
```

```{r}
robust_fn <- .trap_safely(.trap_quietly(function() stop("boom")))
robust_fn()
```

Note the order of composition: `.trap_quietly()` wraps the original
function first, and `.trap_safely()` wraps the result, so that the
error thrown *inside* the quietly-wrapped call is the thing
`.trap_safely()` catches. Composing the other way around --
`.trap_quietly(.trap_safely(.f))` -- would instead capture a function
that never errors in the first place (since `.trap_safely()` already
turned errors into ordinary return values), so there would be nothing
left for the outer `.trap_quietly()` to trap on that front.
