---
title: "minitrap"
---

```{r}
#| include: false
source(file.path("minitrap", "minitrap.R"))
```

The purpose of **minitrap** is to provide zero-dependency versions of two [**purrr**](https://purrr.tidyverse.org/) adverbs that are used to "trap" one or more aspects of the output of another function. Like all adverb functions they take one function as input, and return a modified function as the output. The [README](https://github.com/djnavarro/minis/tree/main/minitrap) provides the full API reference and scope notes; this page provides a short tutorial that shows how to use the functions, and how they compose together.

## Functions

The [**minitrap.R**](https://github.com/djnavarro/minis/blob/main/minitrap/minitrap.R) script supplies two functions:

- `.trap_safely(.f)` is used to wrap `.f` so that a call which would normally produce an error instead returns a list containing two elements named `result` and `error`, with exactly one of the two set to `NULL` depending on whether the unmodified function would have errored when called the same way.
- `.trap_quietly(.f)` is used to wrap `.f` so that all printed output, warnings, and messages are silently captured instead of shown, returning a four-element named list with components `result`, `output`, `warnings`, and `messages`.

It contains no internal helpers.

## `.trap_safely()`

Ordinarily, a function that errors stops execution and requires you to wrap the call in `tryCatch()` if you want to capture the error and take some other action instead of stopping the execution. The goal of `.trap_safely()` is to invert this: it takes a function and returns a new one that never throws an error, instead capturing the error as data. Calling `.trap_safely(log)` on a non-numeric input, for example, doesn't let the resulting error reach the console: instead, it returns a list with `result = NULL` and `error` set to the error condition object that would have been signalled:

```{r}
safe_log <- .trap_safely(log)
safe_log("a")
```

A successful call returns a list with the same two elements, with `error` set to `NULL`
instead, and `result` set to the calculated logarithm. 

```{r}
safe_log(10)
```

This is particularly useful when mapping a function over many inputs and some of them are expected to fail -- rather than the whole iteration aborting on the first error, each element's outcome (success or failure) is captured individually. Combining `.trap_safely()` with [**minimap**](../minimap/)'s `.iter_map()` is a convenient way to illustrate exactly that pattern.

```{r}
#| include: false
source(file.path("minimap", "minimap.R"))
```

```{r}
mixed_list <- list(4, "a", 9) 
mixed_list |> 
  .iter_map(.trap_safely(sqrt)) |>
  .iter_map_lgl(\(r) is.null(r$error))
```

As one would expect, this output shows that `sqrt(4)` succeeds, `sqrt("a")` errors, and `sqrt(9)` succeeds. The error that would normally be produced at the second step does not stop execution, and instead is captured and recorded. (Note that **minitrap** does not itself depend on **minimap**: this page sources that script to show the two minis operating together.)

## `.trap_quietly()`

Where `.trap_safely()` is about errors, `.trap_quietly()` is about outputs: printed output, return values, warnings, and messages. Rather than letting any of that reach the console, `.trap_quietly()` collects it all and hands it back as part of the return value:

```{r}
quiet_fn <- .trap_quietly(function() {
  message("starting")
  warning("using a default")
  42
})
quiet_fn()
```

Note the shape of the result: `result` holds the ordinary return value (i.e. `42`), `warnings` and `messages` are character vectors with one entry per condition raised (in the order they occurred), and `output` is whatever was printed via `cat()`/`print()`, collapsed into a single string -- empty here, since the original function would not have printed anything directly. 

Crucially, `.trap_quietly()` does *not* catch errors -- a function that
errors still aborts the call, same as if it were called directly. 

```{r}
#| error: true
.trap_quietly(function() stop("boom"))()
```


## Composition

As you can see, the two functions serve complementary purposes: `.trap_safely()` catches errors but not outputs, whereas `.trap_quietly()` catches outputs but not errors. Should you need output-trapping and error-trapping together, you can compose the two functions:

```{r}
robust_fn <- .trap_safely(.trap_quietly(function() stop("boom")))
robust_fn()
```

Pay attention to the order of composition: silence the function first, then make it safe. In the example above `.trap_quietly()` wraps the original function first, and `.trap_safely()` wraps the result, so that the error thrown *inside* the quietly-wrapped call becomes the thing that `.trap_safely()` catches. Composing in the opposite order does not produce the behaviour you most likely want.

