---
title: "minijoin"
---

```{r}
#| include: false
source(file.path("minijoin", "minijoin.R"))
```

The purpose of **minijoin** is to provide a zero-dependency slice of
[**dplyr**](https://dplyr.tidyverse.org/)-style syntax for mutating
joins, all built atop base R's `merge()`. The [README](https://github.com/djnavarro/minis/tree/main/minijoin)
provides the full argument reference; this page compares all four join
functions side by side on two small toy data frames with a partial
overlap, then covers the `by` argument in more detail.

## Functions

The [**minijoin.R**](https://github.com/djnavarro/minis/blob/main/minijoin/minijoin.R)
script supplies four core functions:

- `.join_inner_join(x, y, by, ...)` is used to keep only rows with a
  match in both `x` and `y`.
- `.join_left_join(x, y, by, ...)` is used to keep all rows of `x`,
  filling in `NA` for `y`'s columns where there's no match.
- `.join_right_join(x, y, by, ...)` is used to keep all rows of `y`,
  filling in `NA` for `x`'s columns where there's no match.
- `.join_full_join(x, y, by, ...)` is used to keep all rows of both
  `x` and `y`, filling in `NA` on whichever side has no match.

There is also one internal helper:

- `.join_worker()` is the shared implementation that all four join
  functions delegate to, after fixing the right combination of
  `merge()`'s `all.x`/`all.y` arguments; it also restores `x`'s
  original row order, which `merge()` doesn't guarantee on its own.

## Minimal data frames

```{r}
bands <- data.frame(
  band = c("Beatles", "Rolling Stones", "Kinks"),
  founded = c(1960, 1962, 1963)
)
albums <- data.frame(
  band = c("Beatles", "Kinks", "Queen"),
  album = c("Abbey Road", "Arthur", "A Night at the Opera")
)
bands
albums
```

Here we have `bands` and `albums` as the data sets, sharing
`"Beatles"` and `"Kinks"`. The `"Rolling Stones"` exist only in
`bands`, but `"Queen"` exists only in `albums`. The four join types
differ only in which of those non-matching rows they keep.

## Inner join: only rows matching in both

Neither `"Rolling Stones"` nor `"Queen"` appears -- an inner join is the most restrictive of the four, keeping only rows with a counterpart on the other side:

```{r}
.join_inner_join(bands, albums, by = "band")
```

## Left join: all rows of `x` (`bands`)

The `"Rolling Stones"` row is kept, with `NA` for `album`, since it has
no match in `albums`. This is the join to reach for when `x` is your
"main" table and you're enriching it with optional data from `y`:

```{r}
.join_left_join(bands, albums, by = "band")
```

## Right join: all rows of `y` (`albums`)

The `"Queen"` row is kept, with `NA` for `founded`. A right join is
really just a left join with the arguments swapped --
`.join_right_join(x, y, ...)` behaves like `.join_left_join(y, x, ...)`,
with columns in `x`'s original order. Note the row order below:
matched rows still come first, in `bands`' original order, with the
unmatched-in-`x` row (`"Queen"`) appended at the end -- `minijoin`
always restores `x`'s original row order for matched rows, rather than
`y`'s:

```{r}
.join_right_join(bands, albums, by = "band")
```

## Full join: all rows of both

Both `"Rolling Stones"` and `"Queen"` are kept, each with an `NA` in the column they don't have a match for -- a full join never drops a row from either side, so it's the join to reach for when you want to see everything and let missing matches show up as `NA` rather than disappearing silently:

```{r}
.join_full_join(bands, albums, by = "band")
```

## `by` is always explicit

Unlike `dplyr`, there's no auto-detection of shared column names -- `by` must always be supplied. This is a deliberate simplification, not an oversight, to avoid a class of silent bugs from an unintended shared column name (two data frames that happen to both have an `id` column with unrelated meanings, for instance).

When the join column has different names on each side, `by` takes a named character vector instead of a single string, `c("x_name" = "y_name")` -- mirroring `dplyr`'s own syntax for the same situation:

```{r}
albums_renamed <- albums
names(albums_renamed)[1] <- "artist"
albums_renamed

.join_inner_join(bands, albums_renamed, by = c("band" = "artist"))
```

Note that the output keeps `x`'s column name (`band`), not `y`'s (`artist`) -- the join collapses the two into a single column, using whichever name `x` had.
