---
title: "minijoin"
---

```{r}
#| include: false
source(file.path("minijoin", "minijoin.R"))
```

The `minijoin` mini provides a tiny slice of dplyr-like syntax for mutating joins, all built atop base `merge()`. See the [README](https://github.com/djnavarro/minis/tree/main/minijoin) for the full argument reference; this page compares all four side by side on two small toy data frames with a partial overlap, then covers the `by` argument in more detail.

## Functions

The core functions are these:

- `.join_inner_join(x, y, by, ...)` -- keep only rows with a match in
  both `x` and `y`.
- `.join_left_join(x, y, by, ...)` -- keep all rows of `x`, filling in
  `NA` for `y`'s columns where there's no match.
- `.join_right_join(x, y, by, ...)` -- keep all rows of `y`, filling in
  `NA` for `x`'s columns where there's no match.
- `.join_full_join(x, y, by, ...)` -- keep all rows of both `x` and
  `y`, filling in `NA` on whichever side has no match.

Additionally, `minijoin` has one internal helper, `.join_worker()`, which all four join functions delegate to after fixing the right combination of `merge()`'s `all.x`/`all.y` arguments; it also restores `x`'s original
row order, which `merge()` doesn't guarantee on its own.

## Minimal data frames

```{r}
bands <- data.frame(
  band = c("Beatles", "Rolling Stones", "Kinks"),
  founded = c(1960, 1962, 1963)
)
albums <- data.frame(
  band = c("Beatles", "Kinks", "Queen"),
  album = c("Abbey Road", "Arthur", "A Night at the Opera")
)
bands
albums
```

I will be honest. An AI designed these two data frames. Here we have `bands` and `albums`  as the data sets, sharing `"Beatles"` and `"Kinks"`. The `"Rolling Stones"` exist only in `bands`, but `"Queen"` exists only in `albums`. The four join types differ only in which of those non-matching rows they keep. This is perfectly fine as a toy example but -- for reasons I should not need to explain -- I am deeply personally offended by the decision to list *Arthur* rather than *Lola Versus Powerman*, and this is merely the beginning of my long list of concerns. 

## Inner join: only rows matching in both

Neither `"Rolling Stones"` nor `"Queen"` appears -- an inner join is the most restrictive of the four, keeping only rows with a counterpart on the other side:

```{r}
.join_inner_join(bands, albums, by = "band")
```

## Left join: all rows of `x` (`bands`)

`"Rolling Stones"` is kept, with `NA` for `album`, since it has no match in `albums`. This is the join to reach for when `x` is your "main" table and you're enriching it with optional data from `y`:

```{r}
.join_left_join(bands, albums, by = "band")
```

## Right join: all rows of `y` (`albums`)

`"Queen"` is kept, with `NA` for `founded`. A right join is really just a left join with the arguments swapped -- `.join_right_join(x, y, ...)` behaves like `.join_left_join(y, x, ...)`, with columns in `x`'s original order. Note the row order below: matched rows still come first, in `bands`' original order, with the unmatched-in-`x` row (`"Queen"`) appended at the end -- `minijoin` always restores `x`'s original row order for matched rows, rather than `y`'s:

```{r}
.join_right_join(bands, albums, by = "band")
```

## Full join: all rows of both

Both `"Rolling Stones"` and `"Queen"` are kept, each with an `NA` in the column they don't have a match for -- a full join never drops a row from either side, so it's the join to reach for when you want to see everything and let missing matches show up as `NA` rather than disappearing silently:

```{r}
.join_full_join(bands, albums, by = "band")
```

## `by` is always explicit

Unlike `dplyr`, there's no auto-detection of shared column names -- `by` must always be supplied. This is a deliberate simplification, not an oversight, to avoid a class of silent bugs from an unintended shared column name (two data frames that happen to both have an `id` column with unrelated meanings, for instance).

When the join column has different names on each side, `by` takes a named character vector instead of a single string, `c("x_name" = "y_name")` -- mirroring `dplyr`'s own syntax for the same situation:

```{r}
albums_renamed <- albums
names(albums_renamed)[1] <- "artist"
albums_renamed

.join_inner_join(bands, albums_renamed, by = c("band" = "artist"))
```

Note that the output keeps `x`'s column name (`band`), not `y`'s (`artist`) -- the join collapses the two into a single column, using whichever name `x` had.
