---
title: "minijoin"
---

```{r}
#| include: false
source(file.path("minijoin", "minijoin.R"))
```

`minijoin` provides four mutating joins --
`mjoin_inner_join()`/`_left_join()`/`_right_join()`/`_full_join()` --
built on base `merge()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minijoin) for
the full argument reference; this page compares all four side by side
on two small toy data frames with a partial overlap.

```{r}
bands <- data.frame(
  band = c("Beatles", "Rolling Stones", "Kinks"),
  founded = c(1960, 1962, 1963)
)
albums <- data.frame(
  band = c("Beatles", "Kinks", "Queen"),
  album = c("Abbey Road", "Arthur", "A Night at the Opera")
)
bands
albums
```

`bands` and `albums` share `"Beatles"` and `"Kinks"`; `"Rolling
Stones"` exists only in `bands`, and `"Queen"` only in `albums`.

## Inner join: only rows matching in both

```{r}
mjoin_inner_join(bands, albums, by = "band")
```

## Left join: all rows of `x` (`bands`)

`"Rolling Stones"` is kept, with `NA` for `album`:

```{r}
mjoin_left_join(bands, albums, by = "band")
```

## Right join: all rows of `y` (`albums`)

`"Queen"` is kept, with `NA` for `founded`. Note the row order: matched
rows still come first, in `bands`' original order, with the
unmatched-in-`x` row (`"Queen"`) appended at the end -- `minijoin`
always restores `x`'s original row order for matched rows, rather than
`y`'s:

```{r}
mjoin_right_join(bands, albums, by = "band")
```

## Full join: all rows of both

Both `"Rolling Stones"` and `"Queen"` are kept, each with an `NA` in
the column they don't have a match for:

```{r}
mjoin_full_join(bands, albums, by = "band")
```

## `by` is always explicit

Unlike `dplyr`, there's no auto-detection of shared column names --
`by` must always be supplied. This is a deliberate simplification, not
an oversight, to avoid a class of silent bugs from an unintended
shared column name.
