---
title: "minijoin"
---

```{r}
#| include: false
source(file.path("minijoin", "minijoin.R"))
```

`minijoin` provides four mutating joins --
`mjoin_inner_join()`/`_left_join()`/`_right_join()`/`_full_join()` --
built on base `merge()`. See the
[README](https://github.com/djnavarro/minis/tree/main/minijoin) for
the full argument reference; this page compares all four side by side
on two small toy data frames with a partial overlap, then covers the
`by` argument in more detail.

## Functions

- `mjoin_inner_join(x, y, by, ...)` -- keep only rows with a match in
  both `x` and `y`.
- `mjoin_left_join(x, y, by, ...)` -- keep all rows of `x`, filling in
  `NA` for `y`'s columns where there's no match.
- `mjoin_right_join(x, y, by, ...)` -- keep all rows of `y`, filling in
  `NA` for `x`'s columns where there's no match.
- `mjoin_full_join(x, y, by, ...)` -- keep all rows of both `x` and
  `y`, filling in `NA` on whichever side has no match.

`minijoin` has one internal helper, `.mjoin_worker()`, which all four
join functions delegate to after fixing the right combination of
`merge()`'s `all.x`/`all.y` arguments; it also restores `x`'s original
row order, which `merge()` doesn't guarantee on its own.

```{r}
bands <- data.frame(
  band = c("Beatles", "Rolling Stones", "Kinks"),
  founded = c(1960, 1962, 1963)
)
albums <- data.frame(
  band = c("Beatles", "Kinks", "Queen"),
  album = c("Abbey Road", "Arthur", "A Night at the Opera")
)
bands
albums
```

`bands` and `albums` share `"Beatles"` and `"Kinks"`; `"Rolling
Stones"` exists only in `bands`, and `"Queen"` only in `albums`. The
four join types differ only in which of those non-matching rows they
keep.

## Inner join: only rows matching in both

Neither `"Rolling Stones"` nor `"Queen"` appears -- an inner join is
the most restrictive of the four, keeping only rows with a
counterpart on the other side:

```{r}
mjoin_inner_join(bands, albums, by = "band")
```

## Left join: all rows of `x` (`bands`)

`"Rolling Stones"` is kept, with `NA` for `album`, since it has no
match in `albums`. This is the join to reach for when `x` is your
"main" table and you're enriching it with optional data from `y`:

```{r}
mjoin_left_join(bands, albums, by = "band")
```

## Right join: all rows of `y` (`albums`)

`"Queen"` is kept, with `NA` for `founded`. A right join is really
just a left join with the arguments swapped -- `mjoin_right_join(x, y,
...)` behaves like `mjoin_left_join(y, x, ...)`, with columns in `x`'s
original order. Note the row order below: matched rows still come
first, in `bands`' original order, with the unmatched-in-`x` row
(`"Queen"`) appended at the end -- `minijoin` always restores `x`'s
original row order for matched rows, rather than `y`'s:

```{r}
mjoin_right_join(bands, albums, by = "band")
```

## Full join: all rows of both

Both `"Rolling Stones"` and `"Queen"` are kept, each with an `NA` in
the column they don't have a match for -- a full join never drops a
row from either side, so it's the join to reach for when you want to
see everything and let missing matches show up as `NA` rather than
disappearing silently:

```{r}
mjoin_full_join(bands, albums, by = "band")
```

## `by` is always explicit

Unlike `dplyr`, there's no auto-detection of shared column names --
`by` must always be supplied. This is a deliberate simplification, not
an oversight, to avoid a class of silent bugs from an unintended shared
column name (two data frames that happen to both have an `id` column
with unrelated meanings, for instance).

When the join column has different names on each side, `by` takes a
named character vector instead of a single string, `c("x_name" =
"y_name")` -- mirroring `dplyr`'s own syntax for the same situation:

```{r}
albums_renamed <- albums
names(albums_renamed)[1] <- "artist"
albums_renamed

mjoin_inner_join(bands, albums_renamed, by = c("band" = "artist"))
```

Note that the output keeps `x`'s column name (`band`), not `y`'s
(`artist`) -- the join collapses the two into a single column, using
whichever name `x` had.
