---
title: "miniseed"
---

```{r}
#| include: false
source(file.path("miniseed", "miniseed.R"))
```

The purpose of **miniseed** is to provide a zero-dependency
reimplementation of the RNG-seed-management slice of
[**withr**](https://withr.r-lib.org/): `with_seed()`,
`with_preserve_seed()`, `local_seed()`, and `local_preserve_seed()`.
The [README](https://github.com/djnavarro/minis/tree/main/miniseed)
provides scope notes and the full API; this page walks through why
these functions exist, and puts the block-wrapping `with_*` pair and
the deferred-cleanup `local_*` pair side by side so it's clear when
each one is the right shape for the job.

## Functions

The [**miniseed.R**](https://github.com/djnavarro/minis/blob/main/miniseed/miniseed.R)
script supplies four functions:

- `.seed_with_seed(seed, code)` is used to set the RNG seed, evaluate
  `code`, and restore the RNG state as it was found -- so a call to
  `.seed_with_seed()` is reproducible without disturbing the ambient
  RNG stream for whatever runs after it.
- `.seed_with_preserve_seed(code)` is used the same way, but doesn't
  itself call `set.seed()`: it just guarantees that whatever `code`
  does to the RNG state is undone once `code` finishes.
- `.seed_local_seed(seed)` is used to set the RNG seed and schedule the
  restore for when the *calling function* returns, rather than
  wrapping a block of code -- for use directly inside a function body.
- `.seed_local_preserve_seed()` is the `local_*` counterpart to
  `.seed_with_preserve_seed()`: it schedules the restore without
  setting a seed itself.

It's supported by three internal helpers -- there's no need to call
any of them directly. The helpers are these:

- `.seed_get_state()`
- `.seed_set_state()`
- `.seed_defer()`

## Reproducible draws with `.seed_with_seed()`

Ordinarily, two calls to a function like `runif()` give different
results, because each call continues drawing from wherever the
ambient RNG stream happens to be. `.seed_with_seed()` fixes the seed
just for the duration of one expression, which is enough to make that
one call reproducible:

```{r}
identical(.seed_with_seed(42, runif(3)), .seed_with_seed(42, runif(3)))
```

## The ambient stream is left untouched

Crucially, `.seed_with_seed()` doesn't just set a seed and leave it
set -- it restores the RNG state exactly as it found it, so code
running before and after a `.seed_with_seed()` call behaves as though
the call was never there. One way to see this directly is to compare
`.Random.seed` (the object R itself uses to track RNG state) before
and after:

```{r}
set.seed(1)
state_before <- .Random.seed
.seed_with_seed(999, runif(5))
state_after <- .Random.seed
identical(state_before, state_after)
```

The practical consequence is that a draw taken after a
`.seed_with_seed()` call continues the ambient stream exactly as if
that call had been skipped entirely:

```{r}
set.seed(1)
reference_draw <- runif(1)

set.seed(1)
.seed_with_seed(999, runif(3))
draw_after_with_seed <- runif(1)

identical(reference_draw, draw_after_with_seed)
```

## `.seed_with_preserve_seed()`: preserving without seeding

Sometimes the code being wrapped already calls `set.seed()` itself --
a legacy simulation function, say -- and all that's needed is a
guarantee that it doesn't leak its RNG state into whatever runs next.
That's what `.seed_with_preserve_seed()` is for: it restores the RNG
state afterwards, but doesn't set a seed of its own beforehand.

```{r}
legacy_sim <- function() {
  set.seed(123)
  runif(3)
}

set.seed(5)
state_before <- .Random.seed
.seed_with_preserve_seed(legacy_sim())
state_after <- .Random.seed
identical(state_before, state_after)
```

## Side by side: wrapping a block vs. seeding inside a function

Both `with_*` functions take a block of code as an argument, which
works well when the code to seed is a single self-contained
expression. But a function's body is often a sequence of statements
rather than one expression -- and that's exactly the shape
`.seed_local_seed()` is for. It's called directly inside the function,
with no block to wrap, and schedules its cleanup for when that
function returns instead. The following two functions do the same
simulation, one written each way:

```{r}
simulate_with_block <- function(seed) {
  .seed_with_seed(seed, {
    x <- rnorm(3)
    y <- rnorm(3)
    x + y
  })
}

simulate_with_local <- function(seed) {
  .seed_local_seed(seed)
  x <- rnorm(3)
  y <- rnorm(3)
  x + y
}

identical(simulate_with_block(42), simulate_with_local(42))
```

They give identical results, and `.seed_local_seed()` reads a little
more naturally once a function has more than a line or two in its
body -- there's no need to indent the whole thing inside a `{ }` block
just to seed it. The trade-off is that `.seed_local_seed()`'s cleanup
only happens once the *enclosing function* exits, so it's only
reliable when called from inside a function -- calling it at the top
level of a script leaves the seed in place until the session ends (or
another `.seed_*` call resets it).

Like `.seed_with_seed()`, the cleanup still runs even though
`simulate_with_local()` never explicitly restores anything itself, and
it doesn't leak past the call:

```{r}
set.seed(7)
state_before <- .Random.seed
simulate_with_local(42)
state_after <- .Random.seed
identical(state_before, state_after)
```

## `.seed_local_preserve_seed()`

This is the `local_*` counterpart to `.seed_with_preserve_seed()`,
following the same pattern: it schedules the restore for when the
enclosing function returns, without setting a seed itself.

```{r}
wrapped_legacy <- function() {
  .seed_local_preserve_seed()
  legacy_sim()
}

set.seed(5)
state_before <- .Random.seed
wrapped_legacy()
state_after <- .Random.seed
identical(state_before, state_after)
```

## What's deliberately missing

**miniseed** only covers withr's RNG-seed functions. Everything else
in withr -- `with_options()`, `with_dir()`, `with_envvar()`, and so on
-- is a separate concern, unrelated to random number generation, and
would belong in a mini of its own if it's ever needed rather than
being bolted onto this one.
